---
title: "3rd dataset"
author: "Jingyu Liang"
date: "2024-11-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(writexl)
```

# air and wipe datasets (both in log scale)
## air dataset
create the dataset: *personal_diff*
```{r}
personal_diff <- read_excel("neurotoxicant - air 0122_for_code.xlsx", sheet = "personal_entry_adj_difference")
personal_diff <- as.data.frame(personal_diff)
personal_diff <- personal_diff %>% rename(FFID=`SAMPLE ID`)

# change to numeric values
for (i in 1:ncol(personal_diff)){
  personal_diff[,i] <- as.numeric(personal_diff[,i])
}

# change column names
for (i in 2:ncol(personal_diff)) {
  colnames(personal_diff)[i] = paste(colnames(personal_diff)[i], "_air", sep="")
}

# log scale
personal_diff <- bind_cols(
  personal_diff[1], 
  lapply(personal_diff[,2:ncol(personal_diff)], function(x) log(x+1)))
```

delete those not significant elements in air dataset
*significant_major/trace* is in the code 1st dataset
```{r}
# get the significant cols
significant_air <- c("FFID",paste0(significant_major, "_air"), paste0(significant_trace, "_air"))

# delete those not significant elements
personal_diff <- personal_diff[,colnames(personal_diff) %in% significant_air]
```

## wipe dataset log(post)-log(pre)
create the dataset *wipe_diff* that stores the difference of elements in wipe dataset.
```{r}
# difference for major element
major_diff <- major1st_log[,-1] - major2nd_log[,-1]
major_diff <- bind_cols(major1st_log$FFID, major_diff)
colnames(major_diff) <- c("FFID", gsub("1$","", colnames(major_diff)[2:11]))

# difference for trace element
trace_diff <- trace1st_log[,-1] - trace2nd_log[,-1]
trace_diff <- bind_cols(trace1st_log$FFID, trace_diff)
colnames(trace_diff) <- c("FFID", gsub("1$","", colnames(trace_diff)[2:15]))

# merge major diff and trace diff
wipe_diff <- major_diff %>% 
  left_join(trace_diff, by = "FFID")

# change to numeric values
for (i in 1:ncol(wipe_diff)){
  wipe_diff[,i] <- as.numeric(wipe_diff[,i])
}

# change column names
for (i in 2:ncol(wipe_diff)) {
  colnames(wipe_diff)[i] = paste(colnames(wipe_diff)[i], "_wipe", sep="")
}

# change all negative values to 0
wipe_diff[wipe_diff < 0] <- 0
```

delete those not significant elements in wipe dataset
```{r}
# get the significant cols
significant_wipe <- c("FFID",paste0(significant_major, "_wipe"), paste0(significant_trace, "_wipe"))

# delete those not significant elements
wipe_diff <- wipe_diff[,colnames(wipe_diff) %in% significant_wipe]
```


# old code for calculating zeros
```{r}
# make negative values to zero
# personal_diff <- as.data.frame(personal_diff)
# personal_diff[personal_diff < 0 ] <- 0
# wipe_diff <- as.data.frame(wipe_diff)
# wipe_diff[wipe_diff < 0] <- 0

# area_diff <- read_xlsx("neurotoxicant - air 1031.xlsx", sheet = "Area_diff")
# area_diff <- as.data.frame(area_diff)
# area_diff[area_diff < 0] <- 0

# correct the wrong column name
# personal_diff <- personal_diff %>% rename(Co = `Co...4`)
# personal_diff <- personal_diff %>% rename(Co2 = `Co...16`)
# area_diff <- area_diff %>% rename(Co = `Co...4`)
# area_diff <- area_diff %>% rename(Co2 = `Co...16`)

# calculate the percent of zeros
# personal_zeros <- sum(personal_diff == 0, na.rm = TRUE)
# area_zeros <- sum(area_diff == 0, na.rm = TRUE)
# areazero_perc <- area_zeros/175
# perszero_perc <- personal_zeros/450

# write_xlsx(personal_diff, "personal_diff.xlsx")
# write_xlsx(area_diff, "area_diff.xlsx")
```

# correlation table
post dataset for wipe
Note: don't delete non-significant elements
```{r}
wipe_post <- major1st_log %>% left_join(trace1st_log, by = "FFID")

# change column names
colnames(wipe_post) <- c("FFID", gsub("1$","_post", colnames(wipe_post)[2:25]))

# all concentrations: combine air and wipe(post) together
allconcentration <- wipe_post %>% left_join(personal_diff, by = "FFID")
```

correlation table for all
```{r}
corr <- cor(allconcentration[, -1], use = "complete.obs", method = "spearman") 

# Replace NA to 0
corr[is.na(corr)] <- 0

corr <- as.data.frame(corr)

# Melt the correlation matrix into long format
cor_melt <- melt(corr) 
```

## heatmap

### air and wipe(post)
deleted `P_air`, `Cu_air`, `Ni_air`, `Zn_air`, `Pb_air`
deleted `Y_post`, `Se_post`, `Te_post`
```{r}
corr_clean <- corr %>% 
  select(grep("air", names(corr))) %>% 
  slice (grep("post",rownames(corr))) %>%
  select(-`P_air`, -`Cu_air`, -`Ni_air`, -`Zn_air`, -`Pb_air`) %>% # drop some columns
  rownames_to_column(var = "RowName") %>%     # generate a column containing row names
  filter(!RowName %in% c("Y_post", "Se_post", "Te_post")) %>%  # Filter out specific rows
  column_to_rownames(var = "RowName")              # Convert back to row names

wipe_post_vs_air <- pheatmap(corr_clean, display_numbers = TRUE, color = colorRampPalette(c("purple", "white", "orange"))(50))

# ggsave("corr_wipe_post_vs_air.png", plot = wipe_post_vs_air$gtable,  width = 8, height = 5, dpi = 300)
```

### air
Deleted P, Be, Ag, Te
```{r}
corr_air <- corr %>% 
  select(grep("air", names(corr))) %>% 
  slice (grep("air",rownames(corr))) %>%
  select(-`P_air`, -`Be_air`, -`Ag_air`, -`Te_air`) %>%
  rownames_to_column(var = "RowName") %>%          # Convert row names to a column
  filter(!RowName %in% c("P_air", "Be_air", "Ag_air", "Te_air")) %>%  # Filter out specific rows
  column_to_rownames(var = "RowName")              # Convert back to row names

air <- pheatmap(corr_air, display_numbers = TRUE, color = colorRampPalette(c("purple", "white", "orange"))(50))

# ggsave("corr_air.png", plot = air$gtable,  width = 8, height = 5, dpi = 300)
```

### wipe (post)

```{r}
corr_wipe <- corr %>% select(grep("post", names(corr))) %>% slice (grep("post",rownames(corr)))

wipe <- pheatmap(corr_wipe, display_numbers = TRUE, color = colorRampPalette(c("purple", "white", "orange"))(50))

# ggsave("corr_wipe_post.png", plot = wipe$gtable,  width = 8, height = 5, dpi = 300)

```

# stack plot for air dataset
## organize the data
```{r}
# divide by major and trace
type <- c(2, rep(1, 10), rep(0, 14))
personal_diff <- rbind(personal_diff, type)
personal_m <- personal_diff %>% select(1, grep(1 , personal_diff[37,])) %>% 
  slice(-nrow(personal_diff))
personal_t <- personal_diff %>% select(1, grep(0, personal_diff[37,])) %>% 
  slice(-nrow(personal_diff))
personal_diff <- personal_diff[-37,]

# pivot_longer
personal_m_long <- personal_m %>% 
  pivot_longer(2:ncol(personal_m), names_to = "name", values_to = "mg/m3")
personal_m_long$type <- "major"

personal_t_long <- personal_t %>% 
  pivot_longer(2:ncol(personal_t), names_to = "name", values_to = "mg/m3")
personal_t_long$type <- "trace"

# merge major and minor 
air_log <- bind_rows(personal_m_long, personal_t_long)
air_log <- air_log %>% arrange(name)

data <- air_log

# correct the chemical name from Ag_air to Ag
data$name <- gsub("_air$", "", data$name)

# calculate the max for reorder
# data <- data %>%
#   group_by(name) %>%
#   mutate(max_value = max(mg/m3, na.rm = TRUE)) %>%
#   ungroup() %>%
#   arrange(desc(max_value))

# data <- na.omit(data) 

# calculate the sum for each element
# data %<>% group_by(name) %>%
#   mutate(sum = sum(mg/m3, na.rm = TRUE))

data <- data %>% mutate(name = factor(name, levels = unique(name)))
data <- data %>% mutate(type = factor(type, levels = unique(type)))


# calculate the sum for each element post or pre fire and sum of all
sum_personal <- data %>%
  group_by(type,name) %>% 
  summarise(`total_mg/m3` = sum(`mg/m3`, na.rm = TRUE), .groups = "drop") %>%
  ungroup() %>%
  group_by(type) %>%
  mutate(sum = sum(`total_mg/m3`, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(type), desc(`total_mg/m3`))

# factor the name
sum_personal <- sum_personal %>% mutate(name = factor(name, levels = unique(name)))
sum_personal <- sum_personal %>% mutate(type = factor(type, levels = unique(type)))

# mark the location for geom_text
loc_half_sum1 <- c()
majorrows <- grep("major", sum_personal$type)
lastnum <- majorrows[length(majorrows)]
for (i in majorrows ) {
 loc_half_sum1[i] = sum(sum_personal$`total_mg/m3`[i:lastnum])-sum_personal$`total_mg/m3`[i]/2
}

loc_half_sum2 <- c()
tracerows <- grep("trace", sum_personal$type)
lastnumt <- tracerows[length(tracerows)]
for (j in tracerows) {
   loc_half_sum2[j-lastnum] = sum(sum_personal$`total_mg/m3`[j:lastnumt])-sum_personal$`total_mg/m3`[j]/2
}
 
sum_personal %<>% mutate(location = c(loc_half_sum1,loc_half_sum2))

# divide into major and trace
sum_major <- sum_personal %>% slice(grep("major", sum_personal$type))
sum_trace <- sum_personal %>% slice(grep("trace", sum_personal$type))

# write_xlsx(sum_personal, "sum_personal.xlsx")
```


stack bar plot 
for all
```{r}
custom_palette <- c("#8DD3C7", "#FFFFB3", "#BEBADA", "#FB8072", "#80B1D3",
                    "#FDB462", "#B3DE69", "#FCCDE5", "#D9D9D9", "#BC80BD",
                    "#CCEBC5",  "#fb9a99","#1b9e77", "#e31a1c", "#7570b3",
                    "#e7298a", "#66a61e", "#e6ab02", "#a6761d",
                    "#666666", "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c")


plot <- ggplot(sum_personal, aes(x=type, y=`total_mg/m3`, fill=name)) +
  geom_bar(stat = "identity")+
  scale_fill_manual(values = custom_palette) +
  labs(
    title = "Sum of Each Neurotoxicant Concentration (air)",
    x = "Type",
    y = "log(mg/m3)",
    fill = "Neurotoxicants") +
  theme_minimal() +
  theme(legend.position = "none") + 
  geom_text(
     data = sum_personal[c(1:6, 11),],
     aes(x = type, y = location, label = paste(name, round(`total_mg/m3`,2))), 
     color = "black", size = 2.5, vjust = 0)

# write_xlsx (sum_personal[-c(1:6, 11),], "notshowinstack.xlsx")
# custom_legend <- sum_personal[-c(1:6, 11),]
# custom_legend$color <- custom_palette[-c(1:6, 11)]
# custom_legend %<>% select(-type, -sum, -location)
# 
# # Add a y-position column to custom_legend
# custom_legend$y <- seq_len(nrow(custom_legend))  # Generate 1 to 17 for y positions
# 
# # Plot with corrected y mapping
# ggplot() +
#   geom_point(
#     data = custom_legend,
#     aes(x = 1, y = y, color = name),  # Use the y column for proper positioning
#     size = 5, shape = 15  # Square-shaped color patches
#   ) +
#   geom_text(
#     data = custom_legend,
#     aes(
#       x = 1.2, y = y,  # Use the y column for text positioning
#       label = paste(name, "(", round(`total_mg/m3`, 2), "mg/m3)")
#     ),
#     hjust = 0, size = 3
#   ) +
#   scale_color_manual(
#     values = setNames(custom_legend$color, custom_legend$name)
#   ) +
#   theme_void() +
#   theme(legend.position = "none")
# 
# # Combine the main plot and the legend using patchwork or cowplot
# library(patchwork)
# combined_plot <- plot + legend_plot + plot_layout(widths = c(3, 1))
# print(combined_plot)
```

for major
```{r}
ggplot(sum_major, aes(x=type, y=`total_mg/m3`, fill=name)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set3" ) +
  labs(
    title = "Sum of Major Neurotoxicant Concentration (air)",
    x = "Type",
    y = "log(mg/m3)",
    fill = "Neurotoxicants") +
  theme_minimal() +
  theme(legend.position = "none") + 
  geom_text(
     data = sum_major[1:4,],
     aes(x = type, y = location, label = paste(name, round(`total_mg/m3`,2))), 
     color = "black", size = 2.5, vjust = 0)
  
```


stack bar plot 
```{r}
# Generate a palette with 14 colors
custom_palette <- c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", "#e6ab02", "#a6761d",
                    "#666666", "#a6cee3", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c")

ggplot(sum_trace, aes(x=type, y=`total_mg/m3`, fill=name)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_palette) +
  labs(
    title = "Sum of Trace Neurotoxicant Concentration (air)",
    x = "Type",
    y = "log(mg/m3)",
    fill = "Neurotoxicants") +
  theme_minimal() +
  theme(legend.position = "none") + 
  geom_text(
     data = sum_trace[1:7,],
     aes(x = type, y = location, label = paste(name, round(`total_mg/m3`,2))), 
     color = "black", size = 2.5, vjust = 0)
  
```