---
title: "Untitled"
author: "Jingyu Liang"
date: "2026-02-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(reshape2) # for melt()
library(pheatmap)
library(Hmisc) # for rcorr()
library(knitr) # for kable
library(writexl)
library(bkmr)
library(tibble)
library(MASS)
library(glmnet)
library(mice)
```

# Toxic metals Air Dataset
```{r}
air_tox <- read_excel("for code neurotoxicant - air 10312025.xlsx",sheet = "air_tox")

air_tox <- air_tox %>%
  rename(`FFID` = `SAMPLE ID`) %>% 
  mutate(FFID = as.numeric(str_sub(FFID, -2))) %>% 
  arrange(`SAMPLE DATE`) %>% 
  filter(!FFID %in% 79)

# change the sample name (identify Blank row)
air_tox <- air_tox %>%
  mutate(
    is_blank_row=if_else(
      (!is.na(COMMENTS) & str_detect(COMMENTS, regex("Blank", ignore_case = TRUE)) ) , TRUE, FALSE
    )
  ) 

# get the colnames
tox_cols <- names(dplyr::select(air_tox, Ba:Pb))

# ----subtract blank value from the same date----
air_tox <- air_tox %>%
  group_by(`SAMPLE DATE`) %>%
  mutate(
    across(
      all_of(tox_cols),
      ~ {
        den_vec <- .x[is_blank_row]   # extract the blank value for this date

        # If no blank found for this date → return NA
        if (length(den_vec) == 0 || is.na(den_vec[1])) {
          NA_real_
        } else {
          .x - den_vec[1]         # subtract the blank value
        }
      }
    )
  ) %>%
  ungroup() 

air_tox <- air_tox %>%
  filter(!is_blank_row)

# ---- imputation by detection limit; still have 0 ----

# read the detection limit 
mdl <- read_excel("Detection Limit_Air 1031.xlsx",sheet = "Sheet2")
mdl_vec <- setNames(as.numeric(mdl[1,tox_cols]),tox_cols)

# count imputation datapoints of all 0's
impute_0_air_tox <- air_tox %>%
  summarise(across(all_of(tox_cols), ~ sum(.x==0, na.rm=TRUE))) %>%
  rowSums()

# calculate percentage of 0s for each column
zero_pct <- air_tox %>%
  summarise(across(
    all_of(tox_cols),
    ~ mean(.x == 0, na.rm = TRUE) * 100
  )) 

zero_pct_long <- zero_pct %>%
  pivot_longer(
    everything(),
    names_to  = "Variable",
    values_to = "Zero_pct"
  )

# impute 0's with LOD/sqrt(2)
air_tox <- air_tox %>%            # zero values
  mutate(across(all_of(tox_cols), 
                ~ ifelse(.x == 0,
                         mdl_vec[cur_column()]/sqrt(2),
                         .x))) 

# count imputation datapoints of the negative's
impute_negative_air_tox <- air_tox %>%
  summarise(across(all_of(tox_cols), ~ sum(.x<0, na.rm=TRUE))) %>%
  rowSums()

# replace negative values with LOD/sqrt(2)
air_tox <- air_tox %>% 
  mutate(across(all_of(tox_cols), 
                ~ ifelse(.x <0,
                         mdl_vec[cur_column()]/sqrt(2),
                         .x)))

# count imputation datapoints of NA
impute_NA_air_tox <- air_tox %>%
  summarise(across(all_of(tox_cols), ~ sum(is.na(.x)))) %>%
  rowSums()

# impute the LOD/sqrt(2) into NA
air_tox <- air_tox %>% 
  mutate(across(all_of(tox_cols), 
                ~ ifelse(is.na(.x),
                         mdl_vec[cur_column()]/sqrt(2),
                         .x)))

# percentage of whole dataset LOD: only number of zero values
perc_LOD_air_tox <- impute_0_air_tox/
  length(unlist(air_tox[,-c(1:4)]))

# ----divided by volume----
air_tox <- air_tox %>%
  mutate(`VOLUME/DATA`= as.numeric(`VOLUME/DATA`))

# fill in the NA caused by in as.numeric
air_tox[4,2] <- 124.19

# divided by volume and the unit is μg/m3 or ng/l (same magnitude)
air_tox <- air_tox %>%
  mutate(
    across(
      all_of(tox_cols),           
      ~ .x / `VOLUME/DATA`         # divide each column by the column VOLUME/DATA
    )
  ) 

air_tox <- air_tox %>%
  dplyr::select(-c(2:4, 29))


# ----log scale----
air_tox_log <- air_tox %>% mutate(across(all_of(tox_cols), ~log10(.x+0.0001428715))) %>% 
  arrange(FFID)

```

summary table
```{r}

# Summarize all tox columns
summary_table <- air_tox_log %>%
  summarise(across(
    all_of(tox_cols),
    list(
      Mean = ~ mean(.x, na.rm = TRUE),
      SD = ~ sd(.x, na.rm = TRUE),
      Median = ~ median(.x, na.rm = TRUE),
      Min = ~ min(.x, na.rm = TRUE),
      Max = ~ max(.x, na.rm = TRUE)
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  pivot_longer(everything(),
               names_to = c("Variable", ".value"),
               names_sep = "_") %>%
  mutate(`%<LOD of all` = if_else(Variable == first(Variable),
                           sprintf("%.1f%%", perc_LOD_air_tox * 100),
                           ""))  %>%
  # join per-variable zero percentages
  left_join(zero_pct_long, by = "Variable") %>%
  mutate(`%<LOD` = sprintf("%.1f%%", Zero_pct)) %>%
  dplyr::select(`%<LOD of all`, Variable, `%<LOD`, Mean, SD, Median, Min, Max)

# Show table
kable(summary_table, digits = 3, align = 'c')
# write_xlsx(summary_table, "air_tox_summary.xlsx")
```

# Toxic metals Wipe dataset
```{r}
# read into R
wipe_tox <- read_excel("for code neurotoxicant - wipe 11012025.xlsx",sheet = "raw_data_entry")

# reshape the data to get consistent with wipe_PAHs_log
wipe_tox <- wipe_tox %>% 
  pivot_longer(
    cols = -FFID,
    names_to = c("metals","time_code"),
    names_pattern = "([A-Za-z0-9]+)([12])$",
    values_to = "value"
  ) %>% 
  mutate(
    Time = case_when(
      time_code == "1" ~ "post",
      time_code == "2" ~ "pre"
    )
  ) %>% 
  dplyr::select(-time_code) %>% 
  pivot_wider(
    names_from = "metals",
    values_from = "value"
  ) %>% 
  relocate(Time, .before = FFID)

# remove the FFIDs that has single values: 58, 51, 60
wipe_tox <- wipe_tox %>% 
  filter(!FFID %in% c(58, 51, 60))

# ----imputation with detection limit; still have 0----

# read the detection limit in R
mdl <- read_excel("Detection Limit_Air 1031.xlsx", sheet = "LOD_wipe_tox")
mdl_vec <- setNames(as.numeric(mdl[1,tox_cols]),tox_cols)

# count imputation datapoints of NA
impute_NA_wipe_tox <- wipe_tox %>%
  summarise(across(all_of(tox_cols), ~ sum(is.na(.x)))) %>%
  rowSums()

# count imputation datapoints of 0's
impute_0_wipe_tox <- wipe_tox %>%
  summarise(across(all_of(tox_cols), ~ sum(.x==0, na.rm=TRUE))) %>%
  rowSums()

# calculate percentage of 0s for each column
zero_pct <- wipe_tox %>%
  summarise(across(
    all_of(tox_cols),
    ~ mean(.x == 0, na.rm = TRUE) * 100
  )) 

zero_pct_long <- zero_pct %>%
  pivot_longer(
    everything(),
    names_to  = "Variable",
    values_to = "Zero_pct"
  )

# impute 0s & NAs with LOD/sqrt(2)
wipe_tox <- wipe_tox %>%            # zero values
  mutate(across(all_of(tox_cols), 
                ~ ifelse(.x == 0,
                         mdl_vec[cur_column()]/sqrt(2),
                         .x))) %>%  # missing data
  mutate(across(all_of(tox_cols), 
                ~ ifelse(is.na(.x),
                         mdl_vec[cur_column()]/sqrt(2),
                         .x)))

# percentage of LOD: only number of zero values
perc_LOD_wipe_tox <- impute_0_wipe_tox/length(
  unlist(wipe_tox[,-c(1:2)])
)

# ----change units to ppb; still have 0----
# get major and trace cols
major_cols <- names(dplyr::select(wipe_tox,Ba:Zn))
trace_cols <- names(dplyr::select(wipe_tox,Be:Pb))

# change units -> same as sheet "data_entry_adj"
wipe_tox <- wipe_tox %>% 
  mutate(across(all_of(major_cols), ~.x*1000/4.55406)) %>% 
  mutate(across(all_of(trace_cols), ~.x/4.55406)) 

# ----log scale; add a constant----
wipe_tox_log <- wipe_tox %>% 
  mutate(across(all_of(major_cols), ~log10(.x+0.0001428715))) %>% 
  mutate(across(all_of(trace_cols), ~log10(.x+0.0001428715))) 
```

summary table
```{r}

# Summarize all tox columns
summary_table <- wipe_tox_log %>%
  summarise(across(
    all_of(tox_cols),
    list(
      Mean = ~ mean(.x, na.rm = TRUE),
      SD = ~ sd(.x, na.rm = TRUE),
      Median = ~ median(.x, na.rm = TRUE),
      Min = ~ min(.x, na.rm = TRUE),
      Max = ~ max(.x, na.rm = TRUE)
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  pivot_longer(everything(),
               names_to = c("Variable", ".value"),
               names_sep = "_") %>%
  mutate(`%<LOD of all` = if_else(Variable == first(Variable),
                           sprintf("%.1f%%", perc_LOD_wipe_tox * 100),
                           ""))  %>%
  # join per-variable zero percentages
  left_join(zero_pct_long, by = "Variable") %>%
  mutate(`%<LOD` = sprintf("%.1f%%", Zero_pct)) %>%
  dplyr::select(`%<LOD of all`, Variable, `%<LOD`, Mean, SD, Median, Min, Max)

# Show table
# kable(summary_table, digits = 3, align = 'c')
# write_xlsx(summary_table, "wipe_tox_summary.xlsx")
```


## pre & post dataset & Difference
```{r}
wipe_pre_tox <- wipe_tox_log %>% filter(Time=="pre") %>% arrange(FFID)
wipe_post_tox <- wipe_tox_log %>% filter(Time=="post") %>% arrange(FFID)

wipe_diff_tox <- wipe_post_tox
wipe_diff_tox[,3:ncol(wipe_diff_tox)] <- wipe_post_tox[,3:ncol(wipe_post_tox)] - wipe_pre_tox[,3:ncol(wipe_pre_tox)] 

wipe_diff_tox <- wipe_diff_tox[,-1]

wipe_pre_tox <- wipe_pre_tox %>% dplyr::select(-Time)
# count imputation datapoints of the negative's
impute_negative_wipe_tox <- wipe_diff_tox %>%
  summarise(across(all_of(tox_cols), ~ sum(.x<0, na.rm=TRUE))) %>%
  rowSums()

perc_negative_wipe_tox <- impute_negative_wipe_tox/length(
  unlist(wipe_diff_tox[,-1])
)

# change negative values to LOD/sqrt(2)
wipe_diff_tox <- wipe_diff_tox %>%
  mutate(across(all_of(major_cols), 
                ~ ifelse(.x < 0,
                         log10(mdl_vec[cur_column()]/sqrt(2)*1000/4.55406+
                                 0.0001428715),
                         .x ))) %>% 
  mutate(across(all_of(trace_cols), 
                ~ifelse(.x < 0,
                        log10(mdl_vec[cur_column()]/sqrt(2)/4.55406+
                                0.0001428715),
                        .x )))

```

summary table
```{r}

# Summarize all tox columns
summary_table <- wipe_diff_tox %>%
  summarise(across(
    all_of(tox_cols),
    list(
      Mean = ~ mean(.x, na.rm = TRUE),
      SD = ~ sd(.x, na.rm = TRUE),
      Median = ~ median(.x, na.rm = TRUE),
      Min = ~ min(.x, na.rm = TRUE),
      Max = ~ max(.x, na.rm = TRUE)
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  pivot_longer(everything(),
               names_to = c("Variable", ".value"),
               names_sep = "_") %>%
  mutate(`%<LOD` = if_else(Variable == first(Variable),
                           sprintf("%.1f%%", perc_negative_wipe_tox * 100),
                           "")) %>%
  dplyr::select(`%<LOD`, Variable, Mean, SD, Median, Min, Max)

```

## Box plot for log wipe_tox data
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

id_col   <- "FFID"
time_col <- "Time"
chem_cols <- setdiff(names(wipe_tox_log), c(time_col, id_col))

major_set <- c("S","Fe","P","Zn","Ba","Sr","Cu","Mn","Cr","Ni")
trace_set <- setdiff(chem_cols, major_set)

# 1) Paired Wilcoxon (pre vs post) for each chemical + BH
get_p_one <- function(chem) {
  wide <- wipe_tox_log %>%
    dplyr::select(all_of(c(id_col, time_col, chem))) %>%
    mutate(Time = as.character(.data[[time_col]])) %>%
    distinct() %>%
    tidyr::pivot_wider(
      id_cols     = all_of(id_col),
      names_from  = Time,
      values_from = all_of(chem),
      values_fn   = mean
    ) %>%
    filter(!is.na(pre) & !is.na(post))

  n_pair <- nrow(wide)

  pval <- if (n_pair >= 3) {
    suppressWarnings(stats::wilcox.test(wide$post, wide$pre, paired = TRUE, exact = FALSE)$p.value)
  } else {
    NA_real_
  }

  tibble(Chemical = chem, n_pair = n_pair, p_value = pval)
}

p_all <- bind_rows(lapply(chem_cols, get_p_one)) %>%
  mutate(q_value_bh = p.adjust(p_value, method = "BH"))

# Keep only BH-significant chemicals for plotting
sig_chems <- p_all %>%
  filter(!is.na(q_value_bh) & q_value_bh < 0.05) %>%
  pull(Chemical)

# Stars based on BH-adjusted q-values (change to p_value if you want raw p)
star_tbl <- p_all %>%
  mutate(star = case_when(
    is.na(q_value_bh)  ~ "",
    q_value_bh < 0.001 ~ "***",
    q_value_bh < 0.01  ~ "**",
    q_value_bh < 0.05  ~ "*",
    TRUE               ~ ""
  )) %>%
  dplyr::select(Chemical, star)

# 2) Long format for plotting (convert log10 back to original scale)
plot_dat <- wipe_tox_log %>%
  mutate(Time = factor(.data[[time_col]], levels = c("pre", "post"))) %>%
  pivot_longer(all_of(chem_cols), names_to = "Chemical", values_to = "Value_log10") %>%
  mutate(
    Chemical   = as.character(Chemical),
    Value_log10 = as.numeric(Value_log10),
    Value      = 10^Value_log10
  ) %>%
  filter(Chemical %in% sig_chems) %>%
  filter(!is.na(Value), is.finite(Value), Value > 0)

# 3) Order chemicals by max(Value) within major / trace
max_tbl <- plot_dat %>%
  group_by(Chemical) %>%
  summarise(max_value = max(Value, na.rm = TRUE), .groups = "drop")

major_order2 <- max_tbl %>%
  filter(Chemical %in% major_set) %>%
  arrange(desc(max_value)) %>%
  pull(Chemical)

trace_order2 <- max_tbl %>%
  filter(Chemical %in% trace_set) %>%
  arrange(desc(max_value)) %>%
  pull(Chemical)

# 4) Plot settings that match your requirements

y_ticks <- c(1, 10, 100, 1000, 10000, 100000)

y_lab_fun <- function(x) {
  ifelse(x >= 1000, scales::scientific(x), as.character(x))
}

y_top_tick <- 100000
y_star_pos <- y_top_tick + 10

make_boxplot <- function(elements_order, title_text) {

  dat <- plot_dat %>%
    filter(Chemical %in% elements_order) %>%
    mutate(Chemical = factor(Chemical, levels = elements_order))

  if (nrow(dat) == 0) {
    return(ggplot() + theme_void() + labs(title = title_text))
  }

  y_min <- min(dat$Value, na.rm = TRUE)
  y_min <- max(y_min / 1.5, 1e-8)

  y_max <- max(max(dat$Value, na.rm = TRUE), y_star_pos) * 1.05

  ann <- tibble(Chemical = factor(elements_order, levels = elements_order)) %>%
    left_join(star_tbl, by = "Chemical") %>%
    mutate(star = ifelse(is.na(star), "", star),
           y_pos = y_star_pos)

  ggplot(dat, aes(x = Chemical, y = Value, fill = Time)) +
    geom_boxplot(
      na.rm = TRUE,
      position = position_dodge(width = 0.8),
      width = 0.65,
      outlier.shape = 16,
      outlier.size  = 1.6,
      outlier.alpha = 0.8
    ) +
    geom_text(
      data = ann,
      aes(x = Chemical, y = y_pos, label = star),
      inherit.aes = FALSE,
      color = "black",
      size = 5,
      vjust = 0
    ) +
    scale_fill_manual(
      values = c("pre" = "yellow", "post" = "lightblue"),
      name = "Pre/Post-Fire",
      labels = c("pre" = "Pre", "post" = "Post")
    ) +
    scale_y_log10(
      breaks = y_ticks,
      labels = y_lab_fun,
      expand = expansion(mult = c(0.02, 0.06))
    ) +
    coord_cartesian(
      ylim = c(y_min, y_max),
      clip = "off"
    ) +
    labs(
      x = "Toxic metals",
      y = "Parts Per Billion",
      title = title_text
    ) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(colour = "black"),
      plot.margin = margin(t = 25, r = 20, b = 10, l = 10),
      axis.text.x = element_text(hjust = 0.5)
    )
}

# 5) Make plots A (major) and B (trace)
p_major <- make_boxplot(
  major_order2,
  "Pre vs Post concentrations by major elements"
)

p_trace <- make_boxplot(
  trace_order2,
  "Pre vs Post concentrations by trace elements"
)

p_major
p_trace
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

id_col   <- "FFID"
time_col <- "Time"
chem_cols <- setdiff(names(wipe_tox_log), c(time_col, id_col))

major_set <- c("S","Fe","P","Zn","Ba","Sr","Cu","Mn","Cr","Ni")
trace_set <- setdiff(chem_cols, major_set)

# 1) Paired Wilcoxon (pre vs post) for each chemical + BH
get_p_one <- function(chem) {
  wide <- wipe_tox_log %>%
    dplyr::select(all_of(c(id_col, time_col, chem))) %>%
    mutate(Time = as.character(.data[[time_col]])) %>%
    distinct() %>%
    tidyr::pivot_wider(
      id_cols     = all_of(id_col),
      names_from  = Time,
      values_from = all_of(chem),
      values_fn   = mean
    ) %>%
    filter(!is.na(pre) & !is.na(post))

  n_pair <- nrow(wide)

  pval <- if (n_pair >= 3) {
    suppressWarnings(stats::wilcox.test(wide$post, wide$pre, paired = TRUE, exact = FALSE)$p.value)
  } else {
    NA_real_
  }

  tibble(Chemical = chem, n_pair = n_pair, p_value = pval)
}

p_all <- bind_rows(lapply(chem_cols, get_p_one)) %>%
  mutate(q_value_bh = p.adjust(p_value, method = "BH"))

# Keep only BH-significant chemicals
sig_chems <- p_all %>%
  filter(!is.na(q_value_bh) & q_value_bh < 0.05) %>%
  pull(Chemical)

# Stars based on BH-adjusted q-values
star_tbl <- p_all %>%
  mutate(star = case_when(
    is.na(q_value_bh)  ~ "",
    q_value_bh < 0.001 ~ "***",
    q_value_bh < 0.01  ~ "**",
    q_value_bh < 0.05  ~ "*",
    TRUE               ~ ""
  )) %>%
  dplyr::select(Chemical, star)

# 2) Long format for plotting (convert log10 back)
plot_dat <- wipe_tox_log %>%
  mutate(Time = factor(.data[[time_col]], levels = c("pre", "post"))) %>%
  pivot_longer(all_of(chem_cols), names_to = "Chemical", values_to = "Value_log10") %>%
  mutate(
    Chemical    = as.character(Chemical),
    Value_log10 = as.numeric(Value_log10),
    Value       = 10^Value_log10
  ) %>%
  filter(Chemical %in% sig_chems) %>%
  filter(!is.na(Value), is.finite(Value), Value > 0)

# 3) Order chemicals by max(Value) within major / trace
max_tbl <- plot_dat %>%
  group_by(Chemical) %>%
  summarise(max_value = max(Value, na.rm = TRUE), .groups = "drop")

major_order2 <- max_tbl %>%
  filter(Chemical %in% major_set) %>%
  arrange(desc(max_value)) %>%
  pull(Chemical)

trace_order2 <- max_tbl %>%
  filter(Chemical %in% trace_set) %>%
  arrange(desc(max_value)) %>%
  pull(Chemical)

# 4) ONE unified y-axis range for BOTH A and B
y_ticks <- c(1, 10, 100, 1000, 10000, 100000)
y_lab_fun <- function(x) ifelse(x >= 1000, scales::scientific(x), as.character(x))

# global min/max across ALL data that will be plotted (major + trace)
y_min_global <- min(plot_dat$Value, na.rm = TRUE)
y_min_global <- max(y_min_global / 1.5, 1e-12)  # buffer + safety

y_top_tick <- 100000
y_star_pos <- y_top_tick + 10

y_max_data <- max(plot_dat$Value, na.rm = TRUE)
y_max_global <- max(y_max_data, y_star_pos) * 1.08  # headroom so stars are inside

# 5) Plot helper: same ylim for both, stars at 100000+10
make_boxplot <- function(elements_order, title_text) {

  dat <- plot_dat %>%
    filter(Chemical %in% elements_order) %>%
    mutate(Chemical = factor(Chemical, levels = elements_order))

  if (nrow(dat) == 0) {
    return(ggplot() + theme_void() + labs(title = title_text))
  }

  ann <- tibble(Chemical = factor(elements_order, levels = elements_order)) %>%
    left_join(star_tbl, by = "Chemical") %>%
    mutate(
      star = ifelse(is.na(star), "", star),
      y_pos = y_star_pos
    )

  ggplot(dat, aes(x = Chemical, y = Value, fill = Time)) +
    geom_boxplot(
      na.rm = TRUE,
      position = position_dodge(width = 0.8),
      width = 0.65,
      outlier.shape = 16,
      outlier.size  = 1.6,
      outlier.alpha = 0.8
    ) +
    geom_text(
      data = ann,
      aes(x = Chemical, y = y_pos, label = star),
      inherit.aes = FALSE,
      color = "black",
      size = 5,
      vjust = 0
    ) +
    scale_fill_manual(
      values = c("pre" = "yellow", "post" = "lightblue"),
      name = "Pre/Post-Fire",
      labels = c("pre" = "Pre", "post" = "Post")
    ) +
    scale_y_log10(
      breaks = y_ticks,
      labels = y_lab_fun,
      expand = expansion(mult = c(0.02, 0.02))
    ) +
    coord_cartesian(
      ylim = c(y_min_global, y_max_global),
      clip = "off"
    ) +
    labs(
      x = "Toxic metals",
      y = "Parts Per Billion",
      title = title_text
    ) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(colour = "black"),
      plot.margin = margin(t = 18, r = 18, b = 10, l = 10),
      axis.text.x = element_text(hjust = 0.5),
      legend.position = "right"
    )
}

# 6) A (major) and B (trace)
p_major <- make_boxplot(
  major_order2,
  "Pre vs Post concentrations by major elements"
)

p_trace <- make_boxplot(
  trace_order2,
  "Pre vs Post concentrations by trace elements"
)

p_major
p_trace
```


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

id_col   <- "FFID"
time_col <- "Time"
chem_cols <- setdiff(names(wipe_tox_log), c(time_col, id_col))

major_set <- c("S","Fe","P","Zn","Ba","Sr","Cu","Mn","Cr","Ni")
trace_set <- setdiff(chem_cols, major_set)

# 1) Paired Wilcoxon (pre vs post) for each chemical + BH
get_p_one <- function(chem) {
  wide <- wipe_tox_log %>%
    dplyr::select(all_of(c(id_col, time_col, chem))) %>%
    mutate(Time = as.character(.data[[time_col]])) %>%
    distinct() %>%
    tidyr::pivot_wider(
      id_cols     = all_of(id_col),
      names_from  = Time,
      values_from = all_of(chem),
      values_fn   = mean
    ) %>%
    filter(!is.na(pre) & !is.na(post))

  n_pair <- nrow(wide)
  pval <- if (n_pair >= 3) {
    suppressWarnings(stats::wilcox.test(wide$post, wide$pre, paired = TRUE, exact = FALSE)$p.value)
  } else {
    NA_real_
  }

  tibble(Chemical = chem, n_pair = n_pair, p_value = pval)
}

p_all <- bind_rows(lapply(chem_cols, get_p_one)) %>%
  mutate(q_value_bh = p.adjust(p_value, method = "BH"))

sig_chems <- p_all %>%
  filter(!is.na(q_value_bh) & q_value_bh < 0.05) %>%
  pull(Chemical)

star_tbl <- p_all %>%
  mutate(star = case_when(
    is.na(q_value_bh)  ~ "",
    q_value_bh < 0.001 ~ "***",
    q_value_bh < 0.01  ~ "**",
    q_value_bh < 0.05  ~ "*",
    TRUE               ~ ""
  )) %>%
  dplyr::select(Chemical, star)

# 2) Long format for plotting (convert log10 back)
plot_dat <- wipe_tox_log %>%
  mutate(Time = factor(.data[[time_col]], levels = c("pre", "post"))) %>%
  pivot_longer(all_of(chem_cols), names_to = "Chemical", values_to = "Value_log10") %>%
  mutate(
    Chemical    = as.character(Chemical),
    Value_log10 = as.numeric(Value_log10),
    Value       = 10^Value_log10
  ) %>%
  filter(Chemical %in% sig_chems) %>%
  filter(!is.na(Value), is.finite(Value), Value >= 0)

# 3) Order chemicals by max(Value) within major / trace
max_tbl <- plot_dat %>%
  group_by(Chemical) %>%
  summarise(max_value = max(Value, na.rm = TRUE), .groups = "drop")

major_order2 <- max_tbl %>%
  filter(Chemical %in% major_set) %>%
  arrange(desc(max_value)) %>%
  pull(Chemical)

trace_order2 <- max_tbl %>%
  filter(Chemical %in% trace_set) %>%
  arrange(desc(max_value)) %>%
  pull(Chemical)

# 4) Shifted log axis:
c_shift <- 1/9

y_ticks_raw    <- c(1, 10, 100, 1000, 10000, 100000)
y_breaks_shift <- y_ticks_raw + c_shift
y_lab_fun <- function(x_raw) ifelse(x_raw >= 1000, scales::scientific(x_raw), as.character(x_raw))

plot_dat2 <- plot_dat %>%
  mutate(Value_shift = Value + c_shift)

y_lower_view_shift <- c_shift * 0.75

y_top_tick_raw      <- 100000
y_star_raw          <- y_top_tick_raw + 10
y_star_shift        <- y_star_raw + c_shift
y_upper_view_shift  <- (y_star_raw * 1.06) + c_shift

# 5) Plot helper (no title/subtitle/caption; legend no title)
make_boxplot <- function(elements_order) {

  dat <- plot_dat2 %>%
    filter(Chemical %in% elements_order) %>%
    mutate(Chemical = factor(Chemical, levels = elements_order))

  if (nrow(dat) == 0) {
    return(ggplot() + theme_void())
  }

  ann <- tibble(Chemical = factor(elements_order, levels = elements_order)) %>%
    left_join(star_tbl, by = "Chemical") %>%
    mutate(star = ifelse(is.na(star), "", star))

  ggplot(dat, aes(x = Chemical, y = Value_shift, fill = Time)) +
    geom_boxplot(
      na.rm = TRUE,
      position = position_dodge(width = 0.8),
      width = 0.65,
      outlier.shape = 16,
      outlier.size  = 1.6,
      outlier.alpha = 0.8
    ) +
    geom_text(
      data = ann,
      aes(x = Chemical, y = y_star_shift, label = star),
      inherit.aes = FALSE,
      color = "black",
      size = 5,
      vjust = 0
    ) +
    scale_fill_manual(
      values = c("pre" = "yellow", "post" = "lightblue"),
      labels = c("pre" = "pre", "post" = "post"),
      name = NULL
    ) +
    scale_y_log10(
      breaks = y_breaks_shift,
      labels = y_lab_fun(y_ticks_raw),
      expand = expansion(mult = c(0, 0))
    ) +
    coord_cartesian(
      ylim = c(y_lower_view_shift, y_upper_view_shift),
      clip = "off"
    ) +
    labs(
      x = "Toxic metals",
      y = "Parts Per Billion"
    ) +
    theme_minimal() +
    theme(
      legend.title = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(colour = "black"),
      plot.margin = margin(t = 10, r = 18, b = 10, l = 10),
      axis.text.x = element_text(hjust = 0.5)
    )
}

p_major <- make_boxplot(major_order2)
p_trace <- make_boxplot(trace_order2)

p_major
p_trace
```


```{r}
library(dplyr)
library(tidyr)


median_tbl <- wipe_tox_log %>%
  mutate(Time = factor(.data[[time_col]], levels = c("pre", "post"))) %>%
  pivot_longer(all_of(chem_cols), names_to = "Chemical", values_to = "Value_log10") %>%
  filter(Chemical %in% sig_chems) %>%
  filter(!is.na(Value_log10)) %>%
  mutate(
    Value_log10 = as.numeric(Value_log10),
    Value = 10^Value_log10
  ) %>%
  group_by(Chemical, Time) %>%
  summarise(
    n = dplyr::n(),
    median_value = median(Value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Chemical, Time)

median_tbl

median_tbl_wide <- median_tbl %>%
  dplyr::select(Chemical, Time, median_value) %>%
  pivot_wider(names_from = Time, values_from = median_value)

median_tbl_wide
```

# Models Data
## tox: tox_lr
```{r}
# ---- add sum cols for major and trace ----
# major_air
air_tox_log$major_tox_a <- rowSums(air_tox_log[major_cols], na.rm = TRUE)
air_tox_log$trace_tox_a <- rowSums(air_tox_log[trace_cols], na.rm = TRUE)
wipe_diff_tox$major_tox_w <- rowSums(wipe_diff_tox[major_cols], na.rm = TRUE)
wipe_diff_tox$trace_tox_w <- rowSums(wipe_diff_tox[trace_cols], na.rm = TRUE)
wipe_pre_tox$major_tox_w <- rowSums(wipe_pre_tox[major_cols], na.rm = TRUE)
wipe_pre_tox$trace_tox_w <- rowSums(wipe_pre_tox[trace_cols], na.rm = TRUE)

# ---- mark "_wipe" & "_air" & summation----

# mark column names as _wipe
names(wipe_diff_tox)[2:(ncol(wipe_diff_tox)-2)] <- paste0(
  names(wipe_diff_tox)[2:(ncol(wipe_diff_tox)-2)],"_wipe"
)
names(wipe_pre_tox)[2:(ncol(wipe_pre_tox)-2)] <- paste0(
  names(wipe_pre_tox)[2:(ncol(wipe_pre_tox)-2)],"_wipe"
)

# mark the air columns
names(air_tox_log)[2:(ncol(air_tox_log)-2)] <- paste0(
  names(air_tox_log)[2:(ncol(air_tox_log)-2)],"_air"
)

# form the dataset for linear regression
tox_wipe_lr <- wipe_diff_tox %>%
  left_join(mh, by = "FFID")
# tox_wipe_lr <- wipe_pre_tox %>% 
#   left_join(mh, by = "FFID") # !!or wipe_diff_tox!!

tox_air_lr <- air_tox_log %>% 
  left_join(mh, by = "FFID")

# get the column names of all tox (wipe and air)
# tox_wipe_air_cols <- names(tox_lr)[c(2:49)]

# sum
tox_air_lr$sum_air_tox <- rowSums(tox_air_lr[,grep("_air$",names(tox_air_lr))], na.rm = TRUE)
tox_wipe_lr$sum_wipe_tox <- rowSums(tox_wipe_lr[,grep("_wipe$",names(tox_wipe_lr))], na.rm = TRUE)

```

## wipe PAHs & tox: data_lmm
```{r}
data_lmm <- wipe_PAHs_log %>% 
  left_join(mh, by = "FFID") %>% 
  left_join(wipe_tox_log, by = c("FFID", "Time")) 
```


# Corr table for air and wipe-diff
```{r}
air_metal_data <- tox_air_lr %>%
  dplyr::select(-c(sum_air_tox,major_tox_a,trace_tox_a)) %>% 
  dplyr::select(-all_of(26:36))

air_metal_data <- na.omit(air_metal_data)

wipe_metal_data <- wipe_diff_tox %>%
  dplyr::select(-c(major_tox_w,trace_tox_w))

wipe_metal_data <- na.omit(wipe_metal_data)

air_wipepost_metal_data <- air_metal_data %>% left_join(wipe_metal_data, by="FFID") %>% dplyr::select(-c(FFID))
air_wipepost_metal_data <- na.omit(air_wipepost_metal_data)

# Compute correlation matrix 
cor_table <- cor(air_wipepost_metal_data, use = "pairwise.complete.obs", method = "spearman")
mh_vars <- names(air_wipepost_metal_data)[1:24]
toxic_vars <- setdiff(colnames(air_wipepost_metal_data), mh_vars)
cor_mh_tox <- cor_table[toxic_vars,mh_vars]

pheatmap(
  cor_mh_tox,
  display_numbers = TRUE,
  fontsize_number = 9,
  number_color = "black",  
  color = colorRampPalette(c("purple", "white", "orange"))(50),
  main = "Spearman correlations between mental health outcomes and air toxicants"
)

# Do BH correction
# Spearman correlation and p-values
res <- rcorr(as.matrix(air_wipepost_metal_data), type = "spearman")

# Extract unique (upper-triangle) p-values
pvals <- res$P[upper.tri(res$P)]

# Apply BH correction
pvals_adj <- p.adjust(pvals, method = "BH")

# Put adjusted p-values back into a symmetric matrix
P_adj <- matrix(NA, nrow = nrow(res$P), ncol = ncol(res$P))
rownames(P_adj) <- colnames(P_adj) <- colnames(air_wipepost_metal_data)
P_adj[upper.tri(P_adj)] <- pvals_adj
P_adj[lower.tri(P_adj)] <- t(P_adj)[lower.tri(P_adj)]
diag(P_adj) <- 0  # diagonal correlations are 1, no p-value

# 5) Final results
cor_table <- res$r         # Spearman correlations
p_table_adj <- P_adj       # BH-adjusted p-values

p_mh_tox   <- p_table_adj[mh_vars, toxic_vars]

sum(p_mh_tox < 0.05, na.rm = TRUE)

# significance marks (BH-adjusted)
sig_mark <- ifelse(!is.na(p_mh_tox) & p_mh_tox < 0.05, "*", "")

# labels shown inside cells: correlation + star
num_labels <- matrix(
  ifelse(is.na(cor_mh_tox), "",
         sprintf("%.2f%s", cor_mh_tox, sig_mark)),
  nrow = nrow(cor_mh_tox),
  ncol = ncol(cor_mh_tox),
  dimnames = dimnames(cor_mh_tox)
)

pheatmap(
  cor_mh_tox,
  display_numbers = num_labels,
  number_color = "black",
  fontsize_number = 9,
  color = colorRampPalette(c("purple", "white", "orange"))(50)
)
```

# Corr table for all air tox & mh
```{r}
library(dplyr)
library(Hmisc)
library(ggplot2)

air_metal_data <- tox_air_lr %>%
  dplyr::select(-c(sum_air_tox, major_tox_a, trace_tox_a, FFID))

air_metal_data <- na.omit(air_metal_data)

# Compute correlation matrix
cor_table <- cor(air_metal_data, use = "pairwise.complete.obs", method = "spearman")
mh_vars <- c("CESD", "CESD_bi", "ASI", "ASI_s", "ASI_c", "ASI_p",
             "PCL", "FAST", "FAST_bi", "PCL_bi", "ASI_bi")
toxic_vars <- setdiff(colnames(cor_table), mh_vars)
cor_mh_tox <- cor_table[mh_vars, toxic_vars]

# Do BH correction
# Spearman correlation and p-values
res <- rcorr(as.matrix(air_metal_data), type = "spearman")

# Extract unique (upper-triangle) p-values
pvals <- res$P[upper.tri(res$P)]

# Apply BH correction
pvals_adj <- p.adjust(pvals, method = "BH")

# Put adjusted p-values back into a symmetric matrix
P_adj <- matrix(NA, nrow = nrow(res$P), ncol = ncol(res$P))
rownames(P_adj) <- colnames(P_adj) <- colnames(air_metal_data)
P_adj[upper.tri(P_adj)] <- pvals_adj
P_adj[lower.tri(P_adj)] <- t(P_adj)[lower.tri(P_adj)]
diag(P_adj) <- 0

# Final results
cor_table <- res$r
p_table_adj <- P_adj

p_mh_tox <- p_table_adj[mh_vars, toxic_vars]

sum(p_mh_tox < 0.05, na.rm = TRUE)

# labels shown inside cells: correlation only
num_labels <- matrix(
  ifelse(is.na(cor_mh_tox), "",
         sprintf("%.2f", cor_mh_tox)),
  nrow = nrow(cor_mh_tox),
  ncol = ncol(cor_mh_tox),
  dimnames = dimnames(cor_mh_tox)
)

# Convert to plotting data frame
air_plot_df <- as.data.frame(as.table(cor_mh_tox))
colnames(air_plot_df) <- c("Outcome", "Toxicant", "rho")

air_plot_df$p_adj <- as.vector(p_mh_tox)
air_plot_df$significant <- !is.na(air_plot_df$p_adj) & air_plot_df$p_adj < 0.05
air_plot_df$label <- as.vector(num_labels)

ggplot(air_plot_df, aes(x = Toxicant, y = Outcome, fill = rho)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(
    data = subset(air_plot_df, !significant),
    aes(label = label),
    color = "black",
    size = 3.2
  ) +
  geom_text(
    data = subset(air_plot_df, significant),
    aes(label = label),
    color = "black",
    size = 4.2,
    fontface = "bold"
  ) +
  scale_fill_gradient2(
    low = "purple",
    mid = "white",
    high = "orange",
    midpoint = 0,
    name = NULL
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 11),
    axis.text.y = element_text(color = "black", size = 11),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    plot.title = element_text(color = "black", size = 13, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", size = 10)
  )


```

# Corr table for all wipe tox & mh
```{r}
library(dplyr)
library(Hmisc)
library(ggplot2)

wipe_metal_data <- tox_wipe_lr %>%
  dplyr::select(-c(sum_wipe_tox, major_tox_w, trace_tox_w, FFID))

wipe_metal_data <- na.omit(wipe_metal_data)

# Compute correlation matrix
cor_table <- cor(wipe_metal_data, use = "pairwise.complete.obs", method = "spearman")
mh_vars <- c("CESD", "CESD_bi", "ASI", "ASI_s", "ASI_c", "ASI_p",
             "PCL", "FAST", "FAST_bi", "PCL_bi", "ASI_bi")
toxic_vars <- setdiff(colnames(cor_table), mh_vars)
cor_mh_tox <- cor_table[mh_vars, toxic_vars]

# Do BH correction
# Spearman correlation and p-values
res <- rcorr(as.matrix(wipe_metal_data), type = "spearman")

# Extract unique (upper-triangle) p-values
pvals <- res$P[upper.tri(res$P)]

# Apply BH correction
pvals_adj <- p.adjust(pvals, method = "BH")

# Put adjusted p-values back into a symmetric matrix
P_adj <- matrix(NA, nrow = nrow(res$P), ncol = ncol(res$P))
rownames(P_adj) <- colnames(P_adj) <- colnames(wipe_metal_data)
P_adj[upper.tri(P_adj)] <- pvals_adj
P_adj[lower.tri(P_adj)] <- t(P_adj)[lower.tri(P_adj)]
diag(P_adj) <- 0

# Final results
cor_table <- res$r
p_table_adj <- P_adj

p_mh_tox <- p_table_adj[mh_vars, toxic_vars]

sum(p_mh_tox < 0.05, na.rm = TRUE)

# labels shown inside cells: correlation only
num_labels <- matrix(
  ifelse(is.na(cor_mh_tox), "",
         sprintf("%.2f", cor_mh_tox)),
  nrow = nrow(cor_mh_tox),
  ncol = ncol(cor_mh_tox),
  dimnames = dimnames(cor_mh_tox)
)

# Convert to plotting data frame
wipe_plot_df <- as.data.frame(as.table(cor_mh_tox))
colnames(wipe_plot_df) <- c("Outcome", "Toxicant", "rho")

wipe_plot_df$p_adj <- as.vector(p_mh_tox)
wipe_plot_df$significant <- !is.na(wipe_plot_df$p_adj) & wipe_plot_df$p_adj < 0.05
wipe_plot_df$label <- as.vector(num_labels)

ggplot(wipe_plot_df, aes(x = Toxicant, y = Outcome, fill = rho)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(
    data = subset(wipe_plot_df, !significant),
    aes(label = label),
    color = "black",
    size = 3.2
  ) +
  geom_text(
    data = subset(wipe_plot_df, significant),
    aes(label = label),
    color = "black",
    size = 4.2,
    fontface = "bold"
  ) +
  scale_fill_gradient2(
    low = "purple",
    mid = "white",
    high = "orange",
    midpoint = 0,
    name = NULL
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 11),
    axis.text.y = element_text(color = "black", size = 11),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    plot.title = element_text(color = "black", size = 13, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(color = "black", size = 10)
  )


```




# scatter plot of FAST, CESD, PCL, ASI
```{r}
scatter_df <- tox_wipe_lr %>% dplyr::select(FAST, CESD, PCL, ASI)

panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor = 1.2) {
  
  op <- par(usr = c(0, 1, 0, 1))  
  on.exit(par(op))  
  r <- cor(x, y, method = "spearman", use = "complete.obs")  
  txt <- formatC(r, format = "f", digits = digits)  

  text(0.5, 0.5, paste0(prefix, txt), cex = cex.cor)
}

panel.points <- function(x, y, ...) {
  points(x, y, pch = 1, col = "black", ...)  # Custom styling for points
  
  abline(lm(y ~ x), col = "red", lwd = 1)
  
  lines(
    stats::lowess(x, y, f = 0.75),
    col = "blue",
    lwd = 2
  )
}

pairs(
  scatter_df,
  upper.panel = panel.cor,      
  lower.panel = panel.points,   
  diag.panel = NULL,            
  main = "mental health domains"
)

```

```{r}
library(dplyr)

scatter_df <- tox_wipe_lr %>%
  dplyr::select(FAST, CESD, PCL, ASI)

vars <- names(scatter_df)

# 1) Compute all Spearman p-values, then BH-adjust across all unique pairs
pair_list <- combn(vars, 2, simplify = FALSE)

cor_res <- lapply(pair_list, function(v) {
  xname <- v[1]; yname <- v[2]
  x <- scatter_df[[xname]]
  y <- scatter_df[[yname]]
  ok <- complete.cases(x, y)
  n_pair <- sum(ok)

  if (n_pair >= 3) {
    ct <- suppressWarnings(stats::cor.test(x[ok], y[ok], method = "spearman", exact = FALSE))
    data.frame(
      var1 = xname,
      var2 = yname,
      n_pair = n_pair,
      rho = unname(ct$estimate),
      p_value = ct$p.value,
      stringsAsFactors = FALSE
    )
  } else {
    data.frame(
      var1 = xname,
      var2 = yname,
      n_pair = n_pair,
      rho = NA_real_,
      p_value = NA_real_,
      stringsAsFactors = FALSE
    )
  }
})

cor_tbl <- do.call(rbind, cor_res)
cor_tbl$q_value_bh <- p.adjust(cor_tbl$p_value, method = "BH")

# helper to fetch BH q for (x,y) regardless of ordering
get_q <- function(xname, yname, tab) {
  i <- which((tab$var1 == xname & tab$var2 == yname) | (tab$var1 == yname & tab$var2 == xname))
  if (length(i) == 1) tab$q_value_bh[i] else NA_real_
}

# 2) Upper panel: show rho, and add * if BH-significant (q < 0.05)
panel.cor.bh <- function(x, y, digits = 2, cex.cor = 1.2, ...) {
  xname <- deparse(substitute(x))
  yname <- deparse(substitute(y))

  r <- suppressWarnings(cor(x, y, method = "spearman", use = "complete.obs"))
  q <- get_q(xname, yname, cor_tbl)

  star <- if (!is.na(q) && q < 0.05) "*" else ""
  txt <- formatC(r, format = "f", digits = digits)

  op <- par(usr = c(0, 1, 0, 1))
  on.exit(par(op))

  text(0.5, 0.55, txt, cex = cex.cor)
  if (nzchar(star)) text(0.5, 0.35, star, cex = cex.cor * 1.4)
}

# 3) Lower panel: scatter + lm line + lowess smooth (same idea as you had)
panel.points <- function(x, y, ...) {
  points(x, y, pch = 1, col = "black", ...)
  abline(lm(y ~ x), col = "red", lwd = 1)
  lines(stats::lowess(x, y, f = 0.75), col = "blue", lwd = 2)
}

# 4) Draw plot
pairs(
  scatter_df,
  upper.panel = panel.cor.bh,
  lower.panel = panel.points,
  diag.panel = NULL,
  main = "mental health domains (Spearman rho; * indicates BH q < 0.05)"
)
```





# (1) Covariates & LASSO
```{r}
covar <- read_excel("data_questionformodel_onlydemograph 12312025.xlsx", sheet = "data_adj2")
FH_yesno <- read_excel("data_questionformodel_onlyFHyesno 02202026.xlsx", sheet = "data_adj2")

# Convert all columns to numeric using base R
covar[, -ncol(covar)] <- lapply(covar[, -ncol(covar)], as.numeric)
covar_MH <- covar[,c(2:14)]
covar <- covar[,-c(2:14)]

names(covar) <- names(covar) |>
  (\(x) gsub(" ", "_", x, fixed = TRUE))() |>
  (\(x) gsub("'", "_", x, fixed = TRUE))() |>
  (\(x) gsub(",", "_", x, fixed = TRUE))() |>
  (\(x) gsub("-", "_", x, fixed = TRUE))() |>
  (\(x) gsub("–", "_", x, fixed = TRUE))() |>
  (\(x) gsub("—", "_", x, fixed = TRUE))() |>
  (\(x) gsub("/", "_", x, fixed = TRUE))() |>
  (\(x) gsub("\\.+", "_", x))() |>
  (\(x) gsub("[\\(\\)]", "", x))() |>
  (\(x) gsub("__+", "_", x))() |>
  (\(x) gsub("_$", "", x))() |>
  (\(x) gsub("^_", "", x))()


covar_original <- bind_cols(covar, covar_MH, FH_yesno)

# ----add to metals----
tox_air_covar <- tox_air_lr %>% 
  left_join(covar, by = "FFID")

tox_wipe_covar <- tox_wipe_lr %>% 
  left_join(covar, by = "FFID")
```
## Table 2
### two sample t-test
```{r}
library(dplyr)
library(tidyr)
library(purrr)

dat <- covar_original 
outcomes <- c("CESD", "ASI", "FAST", "PCL")

# Helper 1: Welch two-sample t-test for one binary indicator (1 vs 0)
ttest_yes_no <- function(data, indicator, outcome){
  d <- data %>%
    filter(!is.na(.data[[indicator]]), !is.na(.data[[outcome]])) %>%
    filter(.data[[indicator]] %in% c(0,1))

  n0 <- sum(d[[indicator]] == 0)
  n1 <- sum(d[[indicator]] == 1)

  if (n0 < 2 || n1 < 2) {
    return(tibble(
      ref = "0", compare = indicator, outcome = outcome,
      n_ref = n0, n_cat = n1,
      mean_ref = NA_real_, mean_cat = NA_real_,
      diff_cat_minus_ref = NA_real_, t = NA_real_, df = NA_real_, p = NA_real_
    ))
  }

  tt <- t.test(d[[outcome]] ~ d[[indicator]], var.equal = FALSE)

  mean0 <- mean(d[[outcome]][d[[indicator]] == 0])
  mean1 <- mean(d[[outcome]][d[[indicator]] == 1])

  tibble(
    ref = "0", compare = indicator, outcome = outcome,
    n_ref = n0, n_cat = n1,
    mean_ref = mean0, mean_cat = mean1,
    diff_cat_minus_ref = mean1 - mean0,
    t = unname(tt$statistic),
    df = unname(tt$parameter),
    p = tt$p.value
  )
}

# Helper 2: category vs reference using two binary columns (subset to ref==1 OR cat==1)
# Includes a strict mutual-exclusivity filter: drops rows where ref==1 AND cat==1
ttest_cat_vs_ref <- function(data, ref_col, cat_col, outcome){
  d <- data %>%
    filter(!is.na(.data[[ref_col]]), !is.na(.data[[cat_col]]), !is.na(.data[[outcome]])) %>%
    filter(.data[[ref_col]] %in% c(0,1), .data[[cat_col]] %in% c(0,1)) %>%
    filter(.data[[ref_col]] == 1 | .data[[cat_col]] == 1) %>%
    filter(!(.data[[ref_col]] == 1 & .data[[cat_col]] == 1)) %>%   # drop overlaps
    mutate(group = if_else(.data[[cat_col]] == 1, cat_col, ref_col)) %>%
    filter(group %in% c(ref_col, cat_col))

  n_ref <- sum(d$group == ref_col)
  n_cat <- sum(d$group == cat_col)

  if (n_ref < 2 || n_cat < 2) {
    return(tibble(
      ref = ref_col, compare = cat_col, outcome = outcome,
      n_ref = n_ref, n_cat = n_cat,
      mean_ref = NA_real_, mean_cat = NA_real_,
      diff_cat_minus_ref = NA_real_, t = NA_real_, df = NA_real_, p = NA_real_
    ))
  }

  d$group <- factor(d$group, levels = c(ref_col, cat_col))
  tt <- t.test(d[[outcome]] ~ d$group, var.equal = FALSE)

  tibble(
    ref = ref_col, compare = cat_col, outcome = outcome,
    n_ref = n_ref, n_cat = n_cat,
    mean_ref = mean(d[[outcome]][d$group == ref_col]),
    mean_cat = mean(d[[outcome]][d$group == cat_col]),
    diff_cat_minus_ref = mean_cat - mean_ref,
    t = unname(tt$statistic),
    df = unname(tt$parameter),
    p = tt$p.value
  )
}

# ----- 1) Tobacco: Non_user as reference (mutually exclusive) -----
tobacco_ref <- "Non_user"
tobacco_cats <- c("Current_User", "Former_user")

res_tobacco <- expand_grid(cat = tobacco_cats, outcome = outcomes) %>%
  pmap_dfr(\(cat, outcome) ttest_cat_vs_ref(dat, tobacco_ref, cat, outcome)) %>%
  mutate(big_group = "Tobacco User")

# ----- 2) Alcohol: <1_times_per_month as reference (mutually exclusive) -----
alcohol_ref <- "Consume_alcohol_<1_times_per_month"
alcohol_cats <- c("Consume_alcohol_monthly",
                  "Consume_alcohol_weekly",
                  "Consume_alcohol_2_3_times_per_week",
                  "Consume_alcohol_>4_times_per_week")

res_alcohol <- expand_grid(cat = alcohol_cats, outcome = outcomes) %>%
  pmap_dfr(\(cat, outcome) ttest_cat_vs_ref(dat, alcohol_ref, cat, outcome)) %>%
  mutate(big_group = "Consume Alcohol")

# ----- 3) Trauma on duty: Yes vs No for each item (1 vs 0) -----
duty_items <- c("Fire_related_incidents",
                "Natural_disasters_on_duty",
                "Sudden_death_or_loss_of_a_peer_firefighters",
                "Serious_accidents_or_injuries_on_duty",
                "Serious_illness_or_medical_procedures_on_duty")

res_duty <- expand_grid(indicator = duty_items, outcome = outcomes) %>%
  pmap_dfr(\(indicator, outcome) ttest_yes_no(dat, indicator, outcome)) %>%
  mutate(big_group = "Traumatic Events On Duty")

# ----- 4) Trauma childhood: Yes vs No for each item (1 vs 0) -----
child_items <- c("Sudden_death_or_loss_of_a_loved_one",
                 "Natural_disasters_Childhood",
                 "Physical_or_emotional_abuse",
                 "Serious_accidents_or_injuries_in_Childhood",
                 "Bullying_or_peer_victimization",
                 "Serious_illness_or_medical_procedures_in_childhood")

res_child <- expand_grid(indicator = child_items, outcome = outcomes) %>%
  pmap_dfr(\(indicator, outcome) ttest_yes_no(dat, indicator, outcome)) %>%
  mutate(big_group = "Traumatic Events During Childhood")

# ----- 5) Family history: Yes vs No for each item (1 vs 0) -----
fh_items <- c("FH_PTSD", "FH_DEPRESSION", "FH_ANXIETY", "FH_STRESS", "FH_other","FH_yes","FH_no")

res_fh <- expand_grid(indicator = fh_items, outcome = outcomes) %>%
  pmap_dfr(\(indicator, outcome) ttest_yes_no(dat, indicator, outcome)) %>%
  mutate(big_group = "Family History of Mental Health Issues")

# ----- Combine (now all parts share the same schema, so no NA columns) -----
all_results <- bind_rows(
  res_tobacco,
  res_alcohol,
  res_duty,
  res_child,
  res_fh
)

all_results <- all_results %>%
  group_by(big_group) %>%
  mutate(p_BH = p.adjust(p, method = "BH")) %>%
  ungroup()

View(all_results)
```

### Hotelling's t-test
```{r}
library(dplyr)
library(tidyr)
library(purrr)

dat <- covar_original
outcomes <- c("CESD", "ASI", "FAST", "PCL")

# Hotelling's two-sample T^2 (4 outcomes jointly)

hotelling_two_sample <- function(data, group_vec, outcome_cols){
  d <- data %>%
    dplyr::select(all_of(outcome_cols)) %>%
    mutate(.g = group_vec) %>%
    filter(!is.na(.g)) %>%
    filter(.g %in% c(0,1)) %>%
    filter(complete.cases(.))   # require all 4 outcomes observed

  n0 <- sum(d$.g == 0)
  n1 <- sum(d$.g == 1)
  p <- length(outcome_cols)

  # Need enough sample size and invertible pooled covariance
  if (n0 <= p || n1 <= p) {
    return(tibble(
      n0 = n0, n1 = n1, p = p,
      T2 = NA_real_, F = NA_real_, df1 = NA_real_, df2 = NA_real_, p_value = NA_real_
    ))
  }

  X0 <- as.matrix(d[d$.g == 0, outcome_cols, drop = FALSE])
  X1 <- as.matrix(d[d$.g == 1, outcome_cols, drop = FALSE])

  m0 <- colMeans(X0); m1 <- colMeans(X1)
  S0 <- stats::cov(X0); S1 <- stats::cov(X1)

  Sp <- ((n0 - 1) * S0 + (n1 - 1) * S1) / (n0 + n1 - 2)

  # If Sp is singular, cannot compute standard Hotelling T^2
  if (qr(Sp)$rank < p) {
    return(tibble(
      n0 = n0, n1 = n1, p = p,
      T2 = NA_real_, F = NA_real_, df1 = NA_real_, df2 = NA_real_, p_value = NA_real_
    ))
  }

  dbar <- matrix(m1 - m0, ncol = 1)

  T2 <- as.numeric((n0 * n1) / (n0 + n1) * t(dbar) %*% solve(Sp) %*% dbar)

  df1 <- p
  df2 <- n0 + n1 - p - 1
  Fstat <- as.numeric((df2 * T2) / (df1 * (n0 + n1 - 2)))

  pval <- stats::pf(Fstat, df1 = df1, df2 = df2, lower.tail = FALSE)

  tibble(
    n0 = n0, n1 = n1, p = p,
    T2 = T2, F = Fstat, df1 = df1, df2 = df2, p_value = pval
  )
}

hotelling_yes_no <- function(data, indicator, outcome_cols){
  g <- data[[indicator]]
  res <- hotelling_two_sample(data, g, outcome_cols)
  res %>%
    mutate(ref = "0", compare = indicator,
           big_group = NA_character_)
}

hotelling_cat_vs_ref <- function(data, ref_col, cat_col, outcome_cols){
  dsub <- data %>%
    filter(!is.na(.data[[ref_col]]), !is.na(.data[[cat_col]])) %>%
    filter(.data[[ref_col]] %in% c(0,1), .data[[cat_col]] %in% c(0,1)) %>%
    filter(.data[[ref_col]] == 1 | .data[[cat_col]] == 1) %>%
    filter(!(.data[[ref_col]] == 1 & .data[[cat_col]] == 1)) %>%
    mutate(.g = if_else(.data[[cat_col]] == 1, 1L, 0L))  # 0=ref, 1=cat

  res <- hotelling_two_sample(dsub, dsub$.g, outcome_cols)
  res %>%
    mutate(ref = ref_col, compare = cat_col,
           big_group = NA_character_)
}


# 1) Tobacco: Non_user as reference
tobacco_ref <- "Non_user"
tobacco_cats <- c("Current_User", "Former_user")

res_hot_tobacco <- expand_grid(cat = tobacco_cats) %>%
  pmap_dfr(\(cat) hotelling_cat_vs_ref(dat, tobacco_ref, cat, outcomes)) %>%
  mutate(big_group = "Tobacco User")

# 2) Alcohol: <1_times_per_month as reference
alcohol_ref <- "Consume_alcohol_<1_times_per_month"
alcohol_cats <- c("Consume_alcohol_monthly",
                  "Consume_alcohol_weekly",
                  "Consume_alcohol_2_3_times_per_week",
                  "Consume_alcohol_>4_times_per_week")

res_hot_alcohol <- expand_grid(cat = alcohol_cats) %>%
  pmap_dfr(\(cat) hotelling_cat_vs_ref(dat, alcohol_ref, cat, outcomes)) %>%
  mutate(big_group = "Consume Alcohol")

# 3) Trauma on duty: Yes vs No
duty_items <- c("Fire_related_incidents",
                "Natural_disasters_on_duty",
                "Sudden_death_or_loss_of_a_peer_firefighters",
                "Serious_accidents_or_injuries_on_duty",
                "Serious_illness_or_medical_procedures_on_duty")

res_hot_duty <- expand_grid(indicator = duty_items) %>%
  pmap_dfr(\(indicator) hotelling_yes_no(dat, indicator, outcomes) %>% dplyr::select(-big_group)) %>%
  mutate(big_group = "Traumatic Events On Duty")

# 4) Trauma childhood: Yes vs No
child_items <- c("Sudden_death_or_loss_of_a_loved_one",
                 "Natural_disasters_Childhood",
                 "Physical_or_emotional_abuse",
                 "Serious_accidents_or_injuries_in_Childhood",
                 "Bullying_or_peer_victimization",
                 "Serious_illness_or_medical_procedures_in_childhood")

res_hot_child <- expand_grid(indicator = child_items) %>%
  pmap_dfr(\(indicator) hotelling_yes_no(dat, indicator, outcomes) %>% dplyr::select(-big_group)) %>%
  mutate(big_group = "Traumatic Events During Childhood")

# 5) Family history: Yes vs No
fh_items <- c("FH_PTSD", "FH_DEPRESSION", "FH_ANXIETY", "FH_STRESS", "FH_other","FH_yes","FH_no")

res_hot_fh <- expand_grid(indicator = fh_items) %>%
  pmap_dfr(\(indicator) hotelling_yes_no(dat, indicator, outcomes) %>% dplyr::select(-big_group)) %>%
  mutate(big_group = "Family History of Mental Health Issues")

hotelling_results <- bind_rows(
  res_hot_tobacco,
  res_hot_alcohol,
  res_hot_duty,
  res_hot_child,
  res_hot_fh
) %>%
  group_by(big_group) %>%
  mutate(p_BH = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  arrange(big_group, p_value)

View(hotelling_results)
```

### table 2 add SD 
```{r}
library(dplyr)
library(tidyr)
library(purrr)

dat <- covar_original
outcomes <- c("CESD", "ASI", "FAST", "PCL")

# Set this to match your Table 2 header if needed (e.g., 35)
N_total <- nrow(dat)

fmt_mean_sd <- function(x, digits = 1){
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  if (is.nan(m) || is.na(m)) return(NA_character_)
  paste0(
    formatC(m, format = "f", digits = digits),
    " (", formatC(s, format = "f", digits = digits), ")"
  )
}

summarise_row <- function(data, filter_col, label){
  dsub <- data %>% filter(.data[[filter_col]] == 1)

  N <- nrow(dsub)
  pct <- 100 * N / N_total

  tibble(
    Category = label,
    `N (%)` = sprintf("%d (%.0f%%)", N, pct),
    CESD = fmt_mean_sd(dsub$CESD),
    ASI  = fmt_mean_sd(dsub$ASI),
    FAST = fmt_mean_sd(dsub$FAST),
    PCL5 = fmt_mean_sd(dsub$PCL)
  )
}

# Tobacco (mutually exclusive)
tobacco_rows <- tribble(
  ~col,          ~label,
  "Non_user",    "Non-user",
  "Current_User","Current user",
  "Former_user", "Former user"
)

# Alcohol (mutually exclusive)
alcohol_rows <- tribble(
  ~col,                               ~label,
  "Consume_alcohol_<1_times_per_month","<1 Time Per Month",
  "Consume_alcohol_monthly",          "Monthly",
  "Consume_alcohol_weekly",           "Weekly",
  "Consume_alcohol_2_3_times_per_week","2–3 Times Per Week",
  "Consume_alcohol_>4_times_per_week",">4 Times Per Week"
)

# Trauma (overlap allowed): keep the "Yes (1)" rows only (as your prior table did)
duty_rows <- tribble(
  ~col, ~label,
  "Fire_related_incidents", "Fire-related incidents",
  "Natural_disasters_on_duty", "Natural disasters",
  "Sudden_death_or_loss_of_a_peer_firefighters", "Sudden death/loss of a peer firefighter",
  "Serious_accidents_or_injuries_on_duty", "Serious accidents/injuries",
  "Serious_illness_or_medical_procedures_on_duty", "Serious illness/medical procedures"
)

child_rows <- tribble(
  ~col, ~label,
  "Sudden_death_or_loss_of_a_loved_one", "Sudden death/loss of a loved one",
  "Natural_disasters_Childhood", "Natural disasters",
  "Physical_or_emotional_abuse", "Physical/emotional abuse",
  "Serious_accidents_or_injuries_in_Childhood", "Serious accidents/injuries",
  "Bullying_or_peer_victimization", "Bullying/peer victimization",
  "Serious_illness_or_medical_procedures_in_childhood", "Serious illness/medical procedures"
)

fh_rows <- tribble(
  ~col, ~label,
  "FH_PTSD", "PTSD",
  "FH_DEPRESSION", "Depression",
  "FH_ANXIETY", "Anxiety",
  "FH_STRESS", "Stress",
  "FH_other", "Other",
  "FH_yes", "Family History Issues",
  "FH_no", "No Family History"
)

# -------------------------
# Build table2_like with mean (SD)
# -------------------------
table2_like <- bind_rows(
  tobacco_rows %>% pmap_dfr(\(col, label) summarise_row(dat, col, label)),
  alcohol_rows %>% pmap_dfr(\(col, label) summarise_row(dat, col, label)),
  duty_rows    %>% pmap_dfr(\(col, label) summarise_row(dat, col, label)),
  child_rows   %>% pmap_dfr(\(col, label) summarise_row(dat, col, label)),
  fh_rows      %>% pmap_dfr(\(col, label) summarise_row(dat, col, label))
)

View(table2_like)
write_xlsx(table2_like, "table2_like.xlsx")
```


## air selection threshold 70%
```{r}
dep_vars <- c("CESD", "ASI", "ASI_p", "ASI_c", "ASI_s", "PCL", "FAST")
predictor_cols <- c(40:ncol(tox_air_covar))

lasso_selected_vars_air <- list()
lasso_final_models <- list()

N_BOOT <- 100
SELECTION_THRESHOLD <- 0.70   # >70% frequency

set.seed(123)   # GLOBAL seed for full reproducibility

for (dep in dep_vars) {
  
  # 1. Prepare dataset
  dat <- tox_air_covar[, c(dep, names(tox_air_covar)[predictor_cols])]
  dat <- dat[complete.cases(dat), ]
  
  y_full <- dat[[dep]]
  X_full <- as.matrix(dat[, -1])
  
  p <- ncol(X_full)
  var_names <- colnames(X_full)
  
  selection_count <- setNames(rep(0, p), var_names)
  
  # 2. Stable bootstrap LASSO
  for (b in 1:N_BOOT) {
    
    set.seed(123 + b)  # Deterministic bootstrap sample sequence
    
    # Resample
    idx <- sample(1:nrow(dat), replace = TRUE)
    y <- y_full[idx]
    X <- X_full[idx, ]
    
    # FIXED CV folds for this bootstrap iteration
    set.seed(999 + b)
    foldid <- sample(rep(1:10, length.out = nrow(X)))
    
    # CV LASSO using lambda.min
    cvfit <- cv.glmnet(
      x = X,
      y = y,
      alpha = 0.5,
      nfolds = 10,
      foldid = foldid,
      grouped = FALSE
    )
    
    lam <- cvfit$lambda.min
    coef_lasso <- coef(cvfit, s = lam)
    
    selected <- rownames(coef_lasso)[coef_lasso[, 1] != 0]
    selected <- selected[selected != "(Intercept)"]
    
    selection_count[selected] <- selection_count[selected] + 1
  }
  
  # 3. Selection frequency
  selection_freq <- selection_count / N_BOOT
  
  # 4. Keep variables selected ≥ 70%
  selected_vars <- names(selection_freq[selection_freq >= SELECTION_THRESHOLD])
  
  lasso_selected_vars_air[[dep]] <- selected_vars
  
}

lasso_selected_vars_air
```

## wipe selection threshold 70%
```{r}
dep_vars <- c("CESD", "ASI", "ASI_p", "ASI_c", "ASI_s", "PCL", "FAST")
predictor_cols <- c(40:ncol(tox_wipe_covar))

lasso_selected_vars_wipe <- list()
lasso_final_models <- list()

N_BOOT <- 100
SELECTION_THRESHOLD <- 0.70   # >70% frequency

set.seed(123)   

for (dep in dep_vars) {
  
  # 1. Prepare dataset
  dat <- tox_wipe_covar[, c(dep, names(tox_wipe_covar)[predictor_cols])]
  dat <- dat[complete.cases(dat), ]
  
  y_full <- dat[[dep]]
  X_full <- as.matrix(dat[, -1])
  
  p <- ncol(X_full)
  var_names <- colnames(X_full)
  
  selection_count <- setNames(rep(0, p), var_names)
  
  # 2. Stable bootstrap LASSO
  for (b in 1:N_BOOT) {
    
    set.seed(123 + b)  # Deterministic bootstrap sample sequence
    
    # Resample
    idx <- sample(1:nrow(dat), replace = TRUE)
    y <- y_full[idx]
    X <- X_full[idx, ]
    
    # FIXED CV folds for this bootstrap iteration
    set.seed(999 + b)
    foldid <- sample(rep(1:10, length.out = nrow(X)))
    
    # CV LASSO using lambda.min
    cvfit <- cv.glmnet(
      x = X,
      y = y,
      alpha = 0.5,
      nfolds = 10,
      foldid = foldid,
      grouped = FALSE
    )
    
    lam <- cvfit$lambda.min
    coef_lasso <- coef(cvfit, s = lam)
    
    selected <- rownames(coef_lasso)[coef_lasso[, 1] != 0]
    selected <- selected[selected != "(Intercept)"]
    
    selection_count[selected] <- selection_count[selected] + 1
  }
  
  # 3. Selection frequency
  selection_freq <- selection_count / N_BOOT
  
  # 4. Keep variables selected ≥ 70%
  selected_vars <- names(selection_freq[selection_freq >= SELECTION_THRESHOLD])
  
  lasso_selected_vars_wipe[[dep]] <- selected_vars
  
  # 5. Final OLS
  if (length(selected_vars) > 0) {
    form <- as.formula(paste(dep, "~", paste(selected_vars, collapse = " + ")))
    lasso_final_models[[dep]] <- lm(form, data = dat)
  } else {
    lasso_final_models[[dep]] <- NULL
  }
}

lasso_selected_vars_wipe

```

## figure to present

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

make_select_df <- function(sel_list, dataset_label) {
  tibble::enframe(sel_list, name = "Outcome", value = "Predictor") %>%
    tidyr::unnest_longer(Predictor) %>%
    mutate(
      Dataset  = dataset_label,
      Selected = 1L
    )
}

# Long format: only selected pairs
sel_long <- bind_rows(
  make_select_df(lasso_selected_vars_air,  "Air"),
  make_select_df(lasso_selected_vars_wipe, "Wipe")
)

# Outcome order
outcome_order <- c("CESD", "ASI", "ASI_p", "ASI_c", "ASI_s", "PCL", "FAST")
sel_long <- sel_long %>%
  mutate(Outcome = factor(Outcome, levels = outcome_order))

# All covariates across both datasets (y-axis should include everything)
all_covars <- sel_long %>%
  distinct(Predictor) %>%
  pull(Predictor)

# Full grid so every (Outcome, Covariate, Dataset) exists
grid_full <- tidyr::expand_grid(
  Dataset   = c("Air", "Wipe"),
  Outcome   = factor(outcome_order, levels = outcome_order),
  Predictor = all_covars
) %>%
  left_join(
    sel_long %>% dplyr::select(Dataset, Outcome, Predictor, Selected),
    by = c("Dataset", "Outcome", "Predictor")
  ) %>%
  mutate(Selected = ifelse(is.na(Selected), 0L, Selected))

# Order covariates by how often they were selected overall (Air+Wipe)
pred_order <- grid_full %>%
  group_by(Predictor) %>%
  summarise(n_selected = sum(Selected), .groups = "drop") %>%
  arrange(desc(n_selected), Predictor) %>%
  pull(Predictor)

grid_full <- grid_full %>%
  mutate(
    Predictor = factor(Predictor, levels = pred_order),
    Predictor_label = gsub("_", " ", as.character(Predictor))
  )

fix_label <- function(x) {
  lab <- gsub("_", " ", x)
  lab <- gsub("\\bAvg\\b", "Average", lab)          # Avg -> Average
  lab <- gsub("\\b2 3\\b", "2 to 3", lab)           # 2 3 -> 2 to 3
  lab <- gsub("\\binstructor\\b", "instructor rank", lab)  # instructor -> instructor rank
  lab
}

grid_full <- grid_full %>%
  mutate(Predictor_label = fix_label(as.character(Predictor)))

# Then in ggplot, use Predictor_label on y
plot_pts <- grid_full %>%
  filter(Selected == 1L) %>%
  mutate(
    x_num = as.numeric(Outcome),
    x_jit = ifelse(Dataset == "Air", x_num + 0.14, x_num - 0.14)
  )

ggplot(plot_pts, aes(x = x_jit, y = Predictor_label, color = Dataset)) +
  geom_point(size = 2.6) +
  scale_color_manual(values = c("Air" = "red", "Wipe" = "blue")) +
  scale_x_continuous(breaks = seq_along(outcome_order), labels = outcome_order) +
  labs(x = "Outcome", y = "Covariate",
       title = "Covariates selected by LASSO (Air vs Wipe)") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 0.5))
```




# PCA 
## for air
```{r}
# ---- PCA for air exposures ----
# get the tox names
air_vars <- names(tox_air_lr)[grepl("_air$", names(tox_air_lr))]

# PCA
pca_air <- prcomp(tox_air_lr[, air_vars], center = TRUE, scale. = TRUE)

summary(pca_air)

# View the loadings of the first two principal components sorted by absolute value
loadings_air <- abs(pca_air$rotation[, 1:2])  # PC1 & PC2

# Select the top 5 variables with the largest loadings
top_pc1_air <- sort(loadings_air[, 1], decreasing = TRUE)[1:5]
top_pc2_air <- sort(loadings_air[, 2], decreasing = TRUE)[1:5]

cat("Top 5 chemicals in PC1 (air):\n")
print(names(top_pc1_air))

cat("Top 5 chemicals in PC2 (air):\n")
print(names(top_pc2_air))

```
### Find the loadings of each element
```{r}
## Extract loadings for PC1 and PC2
load_mat <- pca_air$rotation[, 1:2]

## Build table
pc_table <- data.frame(
  element   = rownames(load_mat),

  ## PC1
  PC1_loading     = load_mat[, 1],
  PC1_direction   = ifelse(load_mat[, 1] > 0, "positive", "-"),

  ## PC2
  PC2_loading     = load_mat[, 2],
  PC2_direction   = ifelse(load_mat[, 2] > 0, "positive", "-")
)

## Optional: sort by importance in PC1 or PC2
pc_table_PC1_sorted <- pc_table[order(abs(pc_table$PC1_loading), decreasing = TRUE), ]
pc_table_PC2_sorted <- pc_table[order(abs(pc_table$PC2_loading), decreasing = TRUE), ]

write_xlsx(pc_table_PC2_sorted, "pc_table_PC2_sorted.xlsx")


```


## for wipe
```{r}
# ---- PCA for wipe exposures ----
# get the tox names
wipe_vars <- names(tox_wipe_lr)[grepl("_wipe$", names(tox_wipe_lr))]

# PCA
pca_wipe <- prcomp(tox_wipe_lr[, wipe_vars], center = TRUE, scale. = TRUE)

summary(pca_wipe)

# View the loadings of the first two principal components sorted by absolute value
loadings_wipe <- abs(pca_wipe$rotation[, 1:2])  # PC1 & PC2

# Select the top 5 variables with the largest loadings
top_pc1_wipe <- sort(loadings_wipe[, 1], decreasing = TRUE)[1:5]
top_pc2_wipe <- sort(loadings_wipe[, 2], decreasing = TRUE)[1:5]

cat("Top 5 chemicals in PC1 (wipe):\n")
print(names(top_pc1_wipe))

cat("Top 5 chemicals in PC2 (wipe):\n")
print(names(top_pc2_wipe))

```

### Find the loadings of each element
```{r}
## Extract loadings for PC1 and PC2
load_mat <- pca_wipe$rotation[, 1:2]

## Build table
pc_table <- data.frame(
  element   = rownames(load_mat),

  ## PC1
  PC1_loading     = load_mat[, 1],
  PC1_direction   = ifelse(load_mat[, 1] > 0, "positive", "-"),

  ## PC2
  PC2_loading     = load_mat[, 2],
  PC2_direction   = ifelse(load_mat[, 2] > 0, "positive", "-")
)

## Optional: sort by importance in PC1 or PC2
pc_table_PC1_sorted <- pc_table[order(abs(pc_table$PC1_loading), decreasing = TRUE), ]
pc_table_PC2_sorted <- pc_table[order(abs(pc_table$PC2_loading), decreasing = TRUE), ]

write_xlsx(pc_table_PC1_sorted, "pc_table_PC1_sorted.xlsx")
```

## bar plot for loadings
```{r}
library(dplyr)
library(ggplot2)
library(stringr)

df <- tibble::tribble(
  ~element,  ~PC_loading,
  "Mn_air",  0.473026701,
  "Ba_air",  0.321364732,
  "Cu_air",  0.271572817,
  "Ni_air",  0.261097843,
  "Zn_air",  0.191147683,
  "Cr_air",  0.153386926,
  "P_air",   0.150955406,
  "Fe_air",  0.141033716,
  "Sr_air",  0.126489187,
  "Co_air",  0.061096809,
  "V_air",   0.024007464,
  "Mo_air",  0.010105981
)

plot_df <- df %>%
  arrange(desc(PC_loading)) %>%
  mutate(element = factor(element, levels = element))

ggplot(plot_df, aes(x = element, y = PC_loading)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.3f", PC_loading)), vjust = -0.35, size = 4) +
  labs(x = "Toxicant", y = "PC loading") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 0.5))
```


# PCR + covar
## air
```{r}
mh_domains <- c("CESD", "FAST", "ASI", "ASI_p", "ASI_c", "ASI_s", "PCL")

# Add PC1 and PC2 to tox_air_covar
tox_air_covar$PC1_air <- pca_air$x[, 1]
tox_air_covar$PC2_air <- pca_air$x[, 2]

linreg_results_air <- list()

for (dep in mh_domains) {
  
  cat("Running AIR linear regression for:", dep, "\n")
  
  # Covariates selected by LASSO
  covars <- lasso_selected_vars_air[[dep]]
  covars <- covars[covars != dep]   # safety check
  
  # Build regression formula
  if (length(covars) == 0) {
    formula_str <- paste(dep, "~ PC1_air + PC2_air")
  } else {
    formula_str <- paste(dep, "~ PC1_air + PC2_air +", paste(covars, collapse = " + "))
  }
  
  model_lin <- lm(as.formula(formula_str), data = tox_air_covar)
  linreg_results_air[[dep]] <- model_lin
}


```

## wipe
```{r}
# Add PC1 and PC2 to tox_wipe_covar
tox_wipe_covar$PC1_wipe <- pca_wipe$x[, 1]
tox_wipe_covar$PC2_wipe <- pca_wipe$x[, 2]

linreg_results_wipe <- list()

for (dep in mh_domains) {
  
  cat("Running WIPE linear regression for:", dep, "\n")
  
  covars <- lasso_selected_vars_wipe[[dep]]
  covars <- covars[covars != dep]

  # Build regression formula
  if (length(covars) == 0) {
    formula_str <- paste(dep, "~ PC1_wipe + PC2_wipe")
  } else {
    formula_str <- paste(dep, "~ PC1_wipe + PC2_wipe +", paste(covars, collapse = " + "))
  }
  
  model_lin <- lm(as.formula(formula_str), data = tox_wipe_covar)
  linreg_results_wipe[[dep]] <- model_lin
}



```

## create summary tables: extract_significant
```{r}
extract_significant <- function(model, model_name) {
  
  # summary() coefficients
  coefs <- as.data.frame(summary(model)$coefficients)
  coefs$Predictor <- rownames(coefs)
  rownames(coefs) <- NULL
  
  coefs <- coefs %>%
    dplyr::rename(
      Beta = Estimate,
      p_value  = `Pr(>|t|)`
    ) %>%
    dplyr::select(Predictor, Beta, p_value)
  
  # Remove intercept
  coefs <- coefs[coefs$Predictor != "(Intercept)", ]
  
  # 95% CI
  ci <- as.data.frame(confint(model))
  ci$Predictor <- rownames(ci)
  rownames(ci) <- NULL
  colnames(ci)[1:2] <- c("CI_lower", "CI_upper")
  
  coefs <- merge(coefs, ci, by = "Predictor", all.x = TRUE)
  
  # Significance
  coefs$Significance <- cut(
    coefs$p_value,
    breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
    labels = c("***", "**", "*", " ")
  )
  
  # Keep only significant predictors
  coefs <- coefs[coefs$p_value < 0.05, ]
  
  if (nrow(coefs) == 0) return(NULL)
  
  coefs$Model <- model_name
  
  coefs
}


summary_air <- do.call(
  rbind,
  lapply(names(linreg_results_air), function(dep) {
    extract_significant(linreg_results_air[[dep]], dep)
  })
)

View(summary_air)

summary_wipe <- do.call(
  rbind,
  lapply(names(linreg_results_wipe), function(dep) {
    extract_significant(linreg_results_wipe[[dep]], dep)
  })
)

View(summary_wipe)

write_xlsx(summary_air, "Table 3 air_PCR R output.xlsx")
write_xlsx(summary_wipe, "Table 3 wipe_PCR R output.xlsx")
```

# LR: tox in PCs + covar
## air (only PC1 or PC2)
```{r}
# AIR
chem_pc1_air <- names(top_pc1_air)
chem_pc2_air <- names(top_pc2_air)
chem_air <- c(chem_pc1_air, chem_pc2_air)

linreg_results_air_chem <- list()

for (dep in mh_domains) {
  
  covars <- lasso_selected_vars_air[[dep]]
  covars <- covars[covars != dep]
  
  ## ---- PC1 chemicals only ----
  formula_pc1 <- paste(
    dep, "~",
    paste(c(chem_pc1_air, covars), collapse = " + ")
  )
  
  model_pc1 <- lm(as.formula(formula_pc1), data = tox_air_covar)
  linreg_results_air_chem[[paste0(dep, "_PC1chem")]] <- model_pc1
  
  ## ---- PC2 chemicals only ----
  formula_pc2 <- paste(
    dep, "~",
    paste(c(chem_pc2_air, covars), collapse = " + ")
  )
  
  model_pc2 <- lm(as.formula(formula_pc2), data = tox_air_covar)
  linreg_results_air_chem[[paste0(dep, "_PC2chem")]] <- model_pc2
}

summary_air_chem <- do.call(
  rbind,
  lapply(names(linreg_results_air_chem), function(dep) {
    extract_significant(linreg_results_air_chem[[dep]], dep)
  })
)

View(summary_air_chem)

```

## wipe (only PC1 or PC2)
```{r}
# WIPE
chem_pc1_wipe <- names(top_pc1_wipe)
chem_pc2_wipe <- names(top_pc2_wipe)
chem_wipe <- c(chem_pc1_wipe, chem_pc2_wipe)

linreg_results_wipe_chem <- list()

for (dep in mh_domains) {
  
  covars <- lasso_selected_vars_wipe[[dep]]
  covars <- covars[covars != dep]
  
  ## ---- PC1 chemicals only ----
  formula_pc1 <- paste(
    dep, "~",
    paste(c(chem_pc1_wipe, covars), collapse = " + ")
  )
  
  model_pc1 <- lm(as.formula(formula_pc1), data = tox_wipe_covar)
  linreg_results_wipe_chem[[paste0(dep, "_PC1chem")]] <- model_pc1
  
  ## ---- PC2 chemicals only ----
  formula_pc2 <- paste(
    dep, "~",
    paste(c(chem_pc2_wipe, covars), collapse = " + ")
  )
  
  model_pc2 <- lm(as.formula(formula_pc2), data = tox_wipe_covar)
  linreg_results_wipe_chem[[paste0(dep, "_PC2chem")]] <- model_pc2
}


summary_wipe_chem <- do.call(
  rbind,
  lapply(names(linreg_results_wipe_chem), function(dep) {
    extract_significant(linreg_results_wipe_chem[[dep]], dep)
  })
)

View(summary_wipe_chem)

write_xlsx(summary_air_chem, "Table 3 air_LR_LASSO R output.xlsx")
write_xlsx(summary_wipe_chem, "Table 3 wipe_LR_LASSO R output.xlsx")
```

## air & wipe (both PC1 & PC2)
```{r}
linreg_results_air_chem <- list()

for (dep in mh_domains) {
  
  cat("Running AIR chemical LR for:", dep, "\n")
  
  covars <- lasso_selected_vars_air[[dep]]
  covars <- covars[covars != dep]
  
  predictors <- c(chem_air, covars)
  
  formula_str <- paste(
    dep, "~", paste(predictors, collapse = " + ")
  )
  
  model_lin <- lm(as.formula(formula_str), data = tox_air_covar)
  linreg_results_air_chem[[dep]] <- model_lin
}

linreg_results_wipe_chem <- list()

for (dep in mh_domains) {
  
  cat("Running WIPE chemical LR for:", dep, "\n")
  
  covars <- lasso_selected_vars_wipe[[dep]]
  covars <- covars[covars != dep]
  
  predictors <- c(chem_wipe, covars)
  
  formula_str <- paste(
    dep, "~", paste(predictors, collapse = " + ")
  )
  
  model_lin <- lm(as.formula(formula_str), data = tox_wipe_covar)
  linreg_results_wipe_chem[[dep]] <- model_lin
}

summary_air_chem_both <- do.call(
  rbind,
  lapply(names(linreg_results_air_chem), function(dep) {
    extract_significant(linreg_results_air_chem[[dep]], dep)
  })
)

View(summary_air_chem_both)

summary_wipe_chem_both <- do.call(
  rbind,
  lapply(names(linreg_results_wipe_chem), function(dep) {
    extract_significant(linreg_results_wipe_chem[[dep]], dep)
  })
)

View(summary_wipe_chem_both)

write_xlsx(summary_air_chem_both, "Table 3 air_LR_LASSO_both R output.xlsx")
write_xlsx(summary_wipe_chem_both, "Table 3 wipe_LR_LASSO_both R output.xlsx")
```

## plots for beta & CI
```{r}
library(dplyr)
library(ggplot2)
library(stringr)

# 1) Manually enter the table
df <- tibble::tribble(
  ~Model,  ~Predictor, ~Beta,      ~p_value,   ~CI_lower,   ~CI_upper,  ~Significance,
  "ASI_s", "Ba_air",    3.068712,  0.025807,    0.428758,    5.708666,  "*",
  "ASI_p", "Cr_wipe",  -5.829560,  0.004139,   -9.545210,   -2.113910, "**",
  "CESD",  "V_air",   -11.324700,  0.022655,  -20.736600,   -1.912810,  "*",
  "ASI_p", "Ba_air",   -1.465140,  0.049895,   -2.929460,   -0.000820,  "*",
  "ASI_p", "Cd_air",    1.866010,  0.007598,    0.618017,    3.114003, "**",
  "ASI_p", "Cu_air",    1.732974,  0.024772,    0.269868,    3.196081,  "*",
  "ASI_p", "La_air",   -2.254890,  0.002950,   -3.542390,   -0.967390, "**",
  "ASI_p", "Mn_air",    1.903972,  0.013951,    0.477494,    3.330450,  "*",
  "ASI_p", "Mo_air",   -5.796300,  0.001535,   -8.792250,   -2.800360, "**",
  "ASI_p", "Se_air",   -1.161840,  0.022386,   -2.121300,   -0.202380,  "*",
  "ASI_p", "V_air",     5.557629,  0.001874,    2.598339,    8.516918, "**"
)

# Common plotting prep
df_plot <- df %>%
  dplyr::mutate(
    element = Predictor %>%
      str_replace_all("_air$", " (air)") %>%
      str_replace_all("_wipe$", " (wipe)"),
    Model = factor(Model)
  )

# Order y-axis by Beta from + to -
elem_order <- df_plot %>%
  arrange(desc(Beta)) %>%
  distinct(element, .keep_all = TRUE) %>%
  pull(element)

df_plot <- df_plot %>%
  dplyr::mutate(element = factor(element, levels = rev(elem_order)))

pd <- position_dodge(width = 0.7)

# 2) Forest plot (Beta with 95% CI), color = outcome/model
ggplot(df_plot, aes(x = Beta, y = element, color = Model)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), height = 0.2, position = pd) +
  geom_point(position = pd, size = 2) +
  labs(x = "Beta (95% CI)", y = "Element", color = "Outcome") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

```


# BKMR w/ LASSO

## tox_air_covar using PCA
```{r}
# ---- PCA-selected air toxicants ----
pca_selected_metals_air <- c(
  names(top_pc1_air),   # from PC1
  names(top_pc2_air)  # from PC2
)

# ---- BKMR using PCA-selected exposures ----
mh_domains <- c("CESD", "FAST", "ASI", "ASI_p", "ASI_c", "ASI_s", "PCL")
pip_results_air_pca <- tibble()

for (dep in mh_domains) {
  cat("Fitting BKMR for outcome:", dep, "...\n")
  
  # 1. outcome
  Y <- tox_air_covar[[dep]]

  # 2. exposures: PCA-selected metals
  Z <- tox_air_covar %>%
    dplyr::select(all_of(pca_selected_metals_air)) %>%
    as.matrix()

  # 3. covariates from LASSO
  selected_covars <- lasso_selected_vars_air[[dep]]

  if (length(selected_covars) <= 1) {
    X <- NULL
  } else {
    X <- model.matrix(~ . - 1, data = tox_air_covar[, selected_covars, drop = FALSE])
  }

  # 4. BKMR
  set.seed(202)
  fit_km <- kmbayes(
    y      = Y,
    Z      = Z,
    X      = X,
    iter   = 5000,
    varsel = TRUE,
    verbose = FALSE
  )

  # 5. Extract PIPs for PCA-selected metals
  pips <- ExtractPIPs(fit_km)
  high_pip <- pips %>% filter(PIP > 0.5)

  if (nrow(high_pip) > 0) {
    pip_df <- tibble(
      outcome  = dep,
      exposure = high_pip$variable,
      pip      = high_pip$PIP
    )
    pip_results_air_pca <- bind_rows(pip_results_air_pca, pip_df)
  }

  cat("  >> Finished", dep, "\n\n")
}

# View the combined PIP results
View(pip_results_air_pca)
write_xlsx(pip_results_air_pca, "pip_results_air_pca.xlsx")
```




## tox_wipe_covar using PCA
```{r}
# ---- PCA-selected wipe toxicants ----
pca_selected_metals_wipe <- c(
  names(top_pc1_wipe),  # from PC1
  names(top_pc2_wipe)  # from PC2
)

# ---- BKMR using wipe PCA-selected exposures ----
mh_domains <- c("CESD", "FAST", "ASI", "ASI_p", "ASI_c", "ASI_s", "PCL")
pip_results_wipe_pca <- tibble()

for (dep in mh_domains) {
  cat("Fitting BKMR for outcome:", dep, "...\n")
  
  # 1. outcome variable
  Y <- tox_wipe_covar[[dep]]

  # 2. exposure matrix Z: wipe chemicals from PCA
  Z <- tox_wipe_covar %>%
    dplyr::select(all_of(pca_selected_metals_wipe)) %>%
    as.matrix()

  # 3. covariates selected by LASSO
  selected_covars <- lasso_selected_vars_wipe[[dep]]

  if (length(selected_covars) <= 1) {
    X <- NULL
  } else {
    X <- model.matrix(~ . - 1, data = tox_wipe_covar[, selected_covars, drop = FALSE])
  }

  # 4. Run BKMR
  set.seed(303)
  fit_km <- kmbayes(
    y      = Y,
    Z      = Z,
    X      = X,
    iter   = 5000,
    varsel = TRUE,
    verbose = FALSE
  )

  # 5. Extract PIPs > 0.5
  pips <- ExtractPIPs(fit_km)
  high_pip <- pips %>% filter(PIP > 0.0)

  if (nrow(high_pip) > 0) {
    pip_df <- tibble(
      outcome  = dep,
      exposure = high_pip$variable,
      pip      = high_pip$PIP
    )
    pip_results_wipe_pca <- bind_rows(pip_results_wipe_pca, pip_df)
  }

  cat("  >> Finished", dep, "\n\n")
}

# View all selected exposures
View(pip_results_wipe_pca)
write_xlsx(pip_results_wipe_pca, "pip_results_wipe_pca.xlsx")


```
## Bubble plots for PIP
```{r}
library(dplyr)
library(stringr)
library(ggplot2)


pip_tbl <- tibble::tribble(
  ~outcome, ~exposure,  ~PIP,
  "ASI_p",  "Ba_air",   0.8824,
  "ASI_p",  "Se_air",   0.7756,
  "ASI_p",  "Cu_air",   0.7580,
  "ASI_p",  "Mn_air",   0.7432,
  "ASI_p",  "Mo_air",   0.7372,
  "ASI_p",  "V_air",    0.7168,
  "ASI_p",  "S_air",    0.6844,
  "ASI_p",  "Co_air",   0.6796,
  "ASI_p",  "La_air",   0.6736,
  "ASI_p",  "Cd_air",   0.6724,
  "PCL",    "Ba_air",   0.5740,
  "ASI_s",  "Cd_air",   0.5352,
  "PCL",    "Cd_air",   0.5156,
  "ASI_p",  "Co_wipe",  0.9764,
  "ASI_p",  "Sr_wipe",  0.9756,
  "ASI_p",  "V_wipe",   0.9544,
  "ASI_p",  "Cd_wipe",  0.9468,
  "ASI_p",  "Mn_wipe",  0.8540,
  "ASI_p",  "Cr_wipe",  0.8488,
  "ASI_p",  "Ba_wipe",  0.8248,
  "ASI_p",  "Mo_wipe",  0.8104,
  "ASI_p",  "Cu_wipe",  0.7280,
  "ASI_p",  "Ag_wipe",  0.5688
)


plot_df <- pip_tbl %>%
  mutate(
    medium = case_when(
      str_ends(exposure, "_air")  ~ "air",
      str_ends(exposure, "_wipe") ~ "wipe",
      TRUE ~ NA_character_
    ),
    toxicant = str_remove(exposure, "_(air|wipe)$")
  ) %>%
  filter(!is.na(medium)) %>%
  mutate(
    medium  = factor(medium, levels = c("air", "wipe")),
    outcome = factor(outcome, levels = c("ASI_p", "ASI_s", "PCL"))
  )


tox_group <- plot_df %>%
  distinct(toxicant, medium) %>%
  group_by(toxicant) %>%
  summarise(
    has_air  = any(medium == "air"),
    has_wipe = any(medium == "wipe"),
    .groups = "drop"
  ) %>%
  mutate(group = case_when(
    has_air & has_wipe ~ "overlap",
    has_wipe           ~ "wipe_only",
    has_air            ~ "air_only",
    TRUE               ~ "other"
  ))

tox_rank <- plot_df %>%
  group_by(toxicant) %>%
  summarise(maxP = max(PIP, na.rm = TRUE), .groups = "drop") %>%
  left_join(tox_group, by = "toxicant") %>%
  mutate(group = factor(group, levels = c("overlap", "wipe_only", "air_only"))) %>%
  arrange(group, desc(maxP))

tox_levels <- tox_rank$toxicant

plot_df <- plot_df %>%
  mutate(toxicant = factor(toxicant, levels = tox_levels))


ggplot(plot_df, aes(x = toxicant, y = medium)) +
  geom_point(aes(size = PIP, color = outcome), alpha = 0.9) +
  scale_size_continuous(range = c(1, 15)) +
  scale_color_manual(values = c("ASI_p" = "orange", "ASI_s" = "lightblue", "PCL" = "purple")) +
  labs(x = "Toxicants", y = NULL, size = "PIP", color = NULL) +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    axis.text.y = element_text(size = 16),
    panel.grid.major.y = element_blank()
  )
```

```{r}
library(dplyr)
library(stringr)
library(ggplot2)

pip_tbl <- tibble::tribble(
  ~outcome, ~exposure,  ~PIP,
  "ASI_p",  "Ba_air",   0.8824,
  "ASI_p",  "Se_air",   0.7756,
  "ASI_p",  "Cu_air",   0.7580,
  "ASI_p",  "Mn_air",   0.7432,
  "ASI_p",  "Mo_air",   0.7372,
  "ASI_p",  "V_air",    0.7168,
  "ASI_p",  "S_air",    0.6844,
  "ASI_p",  "Co_air",   0.6796,
  "ASI_p",  "La_air",   0.6736,
  "ASI_p",  "Cd_air",   0.6724,
  "PCL",    "Ba_air",   0.5740,
  "ASI_s",  "Cd_air",   0.5352,
  "PCL",    "Cd_air",   0.5156,
  "ASI_p",  "Co_wipe",  0.9764,
  "ASI_p",  "Sr_wipe",  0.9756,
  "ASI_p",  "V_wipe",   0.9544,
  "ASI_p",  "Cd_wipe",  0.9468,
  "ASI_p",  "Mn_wipe",  0.8540,
  "ASI_p",  "Cr_wipe",  0.8488,
  "ASI_p",  "Ba_wipe",  0.8248,
  "ASI_p",  "Mo_wipe",  0.8104,
  "ASI_p",  "Cu_wipe",  0.7280,
  "ASI_p",  "Ag_wipe",  0.5688
)
plot_df <- pip_tbl %>%
  mutate(
    medium = case_when(
      str_ends(exposure, "_air")  ~ "air",
      str_ends(exposure, "_wipe") ~ "wipe",
      TRUE ~ NA_character_
    ),
    toxicant = str_remove(exposure, "_(air|wipe)$")
  ) %>%
  filter(!is.na(medium)) %>%
  mutate(
    medium  = factor(medium, levels = c("air", "wipe")),
    outcome = factor(outcome, levels = c("ASI_p", "ASI_s", "PCL"))
  )


tox_group <- plot_df %>%
  distinct(toxicant, medium) %>%
  group_by(toxicant) %>%
  summarise(
    has_air  = any(medium == "air"),
    has_wipe = any(medium == "wipe"),
    .groups = "drop"
  ) %>%
  mutate(group = case_when(
    has_air & has_wipe ~ "overlap",
    has_wipe           ~ "wipe_only",
    has_air            ~ "air_only",
    TRUE               ~ "other"
  ))

tox_rank <- plot_df %>%
  group_by(toxicant) %>%
  summarise(maxP = max(PIP, na.rm = TRUE), .groups = "drop") %>%
  left_join(tox_group, by = "toxicant") %>%
  mutate(group = factor(group, levels = c("overlap", "wipe_only", "air_only"))) %>%
  arrange(group, desc(maxP))

tox_levels <- tox_rank$toxicant

plot_df <- plot_df %>%
  mutate(toxicant = factor(toxicant, levels = tox_levels))

ggplot(plot_df, aes(x = toxicant, y = medium)) +
  geom_point(aes(size = PIP, color = outcome), alpha = 0.9) +
  scale_size_continuous(range = c(1, 15)) +
  scale_color_manual(values = c("ASI_p" = "orange", "ASI_s" = "lightblue", "PCL" = "purple")) +
  labs(x = "Toxicants", y = NULL, size = "PIP", color = NULL) +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.text = element_text(size = 16),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 13),
    axis.text.y = element_text(size = 16),
    panel.grid.major.y = element_blank()
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 6)),
    size = guide_legend()
  )
```


# Univariate Exposure-Response analysis
## tox_wipe_covar
PART 1: Compute and store BKMR univariate ER curves
```{r}
# --- PART 1: Compute and store BKMR univariate ER curves ---
univar_ER_wipe <- list()

for (dep in mh_domains) {
  cat("Calculating BKMR univariate ER curves for:", dep, "...\n")
  
  # 1. outcome
  Y <- tox_wipe_covar[[dep]]
  
  # 2. exposure matrix (PCA-selected wipe metals)
  Z <- tox_wipe_covar %>%
    dplyr::select(all_of(pca_selected_metals_wipe)) %>%
    as.matrix()
  
  # 3. covariates
  selected_covars <- lasso_selected_vars_wipe[[dep]]
  if (length(selected_covars) <= 1) {
    X <- NULL
  } else {
    X <- model.matrix(~ . - 1, data = tox_wipe_covar[, selected_covars, drop = FALSE])
  }
  
  # 4. Fit BKMR
  set.seed(303)
  fit_km <- kmbayes(
    y      = Y,
    Z      = Z,
    X      = X,
    iter   = 5000,
    varsel = TRUE,
    verbose = FALSE
  )
  
  # 5. Compute univariate exposure-response curves
  er_curves <- PredictorResponseUnivar(
    fit     = fit_km,
    sel     = 1:ncol(Z),
    method  = "approx"
  )
  
  # Save result for this outcome
  univar_ER_wipe[[dep]] <- er_curves
  
  cat("  >> Finished", dep, "\n\n")
}
```


part 2: Plot ER curves for one mental health domain
```{r}
# ---- Plot ER curves for 7 mental health domains ----

dep_to_plot <- "ASI"
er_obj <- univar_ER_wipe[[dep_to_plot]]
metal_names <- pca_selected_metals_wipe

# Constant used in log10(x + const)
shift_const <- 0.0001428715

for (j in seq_along(metal_names)) {
  
  metal_name <- metal_names[j]
  
  df_plot <- er_obj %>% 
    filter(variable == metal_name)
  
  if (nrow(df_plot) < 2) {
    message("No ER curve for ", metal_name)
    next
  }
  
  df_plot$upper <- df_plot$est + 1.96 * df_plot$se
  df_plot$lower <- df_plot$est - 1.96 * df_plot$se

  # Compute tick marks in log-scale and convert to raw scale
  xticks_log <- pretty(df_plot$z)
  xticks_raw <- round(10^xticks_log - shift_const, 3)

  plot(df_plot$z, df_plot$est, type = "l", lwd = 2,
       xlab = paste0(metal_name, " (ppb)"),
       ylab = paste("Estimated ER for", dep_to_plot),
       main = paste("BKMR Univariate ER Curve:", metal_name),
       xaxt = "n")
  
  axis(side = 1, at = xticks_log, labels = xticks_raw)
  
  lines(df_plot$z, df_plot$upper, lty = 2, col = "gray40")
  lines(df_plot$z, df_plot$lower, lty = 2, col = "gray40")
}


```

part 3: all in one plot
```{r}
# Dependent variables
dep_vars <- c("CESD", "ASI", "ASI_p", "ASI_c", "ASI_s", "PCL", "FAST")

# Constant used in log10(x + const)
shift_const <- 0.0001428715

for (dep in dep_vars) {
  
  cat("Plotting ER curves for:", dep, "\n")
  
  er_obj <- univar_ER_wipe[[dep]]
  
  df_all <- er_obj
  
  # Convert log10(x + c) → raw x value (ppb) for label display
  df_all <- df_all %>%
    mutate(
      z_raw = round(10^z - shift_const, 3)
    )

  p <- ggplot(df_all, aes(x = z, y = est)) +
    geom_line(size = 1) +
    facet_wrap(~ variable, scales = "free_x") +
    labs(
      title = paste("BKMR Univariate Exposure–Response Curves for", dep),
      x = "Exposure (ppb)",  # <-- changed to ppb
      y = paste("Estimated ER for", dep)
    ) +
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "white"),
      plot.title = element_text(hjust = 0.5)
    ) +
    scale_x_continuous(
      breaks = pretty(df_all$z),
      labels = function(z) round(10^z - shift_const, 3)
    )
  
  print(p)
  
 # Save each plot
  ggsave(
    filename = paste0("ER_curve_wipe_", dep, ".png"),
    plot = p,
    width = 10,
    height = 7,
    dpi = 300
  )
  
}


```

## tox_air_covar
PART 1: Compute and store BKMR univariate ER curves
```{r}
# --- PART 1: Compute and store BKMR univariate ER curves ---
univar_ER_air <- list()

for (dep in mh_domains) {
  cat("Calculating BKMR univariate ER curves for:", dep, "...\n")
  
  # 1. outcome
  Y <- tox_air_covar[[dep]]
  
  # 2. exposure matrix (PCA-selected air metals)
  Z <- tox_air_covar %>%
    dplyr::select(all_of(pca_selected_metals_air)) %>%
    as.matrix()
  
  # 3. covariates
  selected_covars <- lasso_selected_vars_air[[dep]]
  if (length(selected_covars) <= 1) {
    X <- NULL
  } else {
    X <- model.matrix(~ . - 1, data = tox_air_covar[, selected_covars, drop = FALSE])
  }
  
  # 4. Fit BKMR
  set.seed(303)
  fit_km <- kmbayes(
    y      = Y,
    Z      = Z,
    X      = X,
    iter   = 5000,
    varsel = TRUE,
    verbose = FALSE
  )
  
  # 5. Compute univariate exposure-response curves
  er_curves <- PredictorResponseUnivar(
    fit     = fit_km,
    sel     = 1:ncol(Z),
    method  = "approx"
  )
  
  # Save result for this outcome
  univar_ER_air[[dep]] <- er_curves
  
  cat("  >> Finished", dep, "\n\n")
}
```

part 2: Plot ER curves for one mental health domain
```{r}
# ---- Plot ER curves for 7 mental health domains ----

dep_to_plot <- "ASI_p"
er_obj <- univar_ER_air[[dep_to_plot]]
metal_names <- pca_selected_metals_air

# Constant used in log10(x + const)
shift_const <- 0.0001428715

for (j in seq_along(metal_names)) {
  
  metal_name <- metal_names[j]
  
  df_plot <- er_obj %>% 
    filter(variable == metal_name)
  
  if (nrow(df_plot) < 2) {
    message("No ER curve for ", metal_name)
    next
  }
  
  df_plot$upper <- df_plot$est + 1.96 * df_plot$se
  df_plot$lower <- df_plot$est - 1.96 * df_plot$se

  # Compute tick marks in log-scale and convert to raw scale
  xticks_log <- pretty(df_plot$z)
  xticks_raw <- round(10^xticks_log - shift_const, 3)

  plot(df_plot$z, df_plot$est, type = "l", lwd = 2,
       xlab = bquote(.(metal_name) ~ "(" * mu * "g/" * m^3 * ")"),
       ylab = paste("Estimated ER for", dep_to_plot),
       main = paste("BKMR Univariate ER Curve:", metal_name),
       xaxt = "n")
  
  axis(side = 1, at = xticks_log, labels = xticks_raw)
  
  lines(df_plot$z, df_plot$upper, lty = 2, col = "gray40")
  lines(df_plot$z, df_plot$lower, lty = 2, col = "gray40")
}


```

part 3: all in one plot
```{r}
# Dependent variables
dep_vars <- c("CESD", "ASI", "ASI_p", "ASI_c", "ASI_s", "PCL", "FAST")

# Constant used in log10(x + const)
shift_const <- 0.0001428715

for (dep in dep_vars) {
  
  cat("Plotting ER curves for:", dep, "\n")
  
  er_obj <- univar_ER_air[[dep]]
  
  df_all <- er_obj
  
  # Convert log10(x + c) → raw x value (ppb) for label display
  df_all <- df_all %>%
    mutate(
      z_raw = round(10^z - shift_const, 3)
    )

  p <- ggplot(df_all, aes(x = z, y = est)) +
    geom_line(size = 1) +
    facet_wrap(~ variable, scales = "free_x") +
    labs(
      title = paste("BKMR Univariate Exposure–Response Curves for", dep),
      x = expression("Exposure (" * mu * "g/" * m^3 * ")"),
      y = paste("Estimated ER for", dep)
    ) +
    theme_bw(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "white"),
      plot.title = element_text(hjust = 0.5)
    ) +
    scale_x_continuous(
      breaks = pretty(df_all$z),
      labels = function(z) round(10^z - shift_const, 3)
    )
  
  print(p)
  
  # Save each plot
  ggsave(
    filename = paste0("ER_curve_air_", dep, ".png"),
    plot = p,
    width = 10,
    height = 7,
    dpi = 300
  )
  
}


```